
---
title: "Lecture 23" 
subtitle: "Spatio-temporal Models"
author: "Colin Rundel"
date: "04/17/2017"
fontsize: 11pt
output: 
  beamer_presentation:
    theme: metropolis
    highlight: pygments
    fig_width: 8
    fig_height: 5
    fig_caption: false
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: ../settings.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning=FALSE, message=FALSE)
options(width=80)

set.seed(20170417)

library(magrittr)
library(ggplot2)
library(tidyr)
library(rjags)
library(gridExtra)
library(purrr)
library(forcats)
library(fields)
library(sf)
library(raster)
library(dplyr)
library(readr)
library(tibble)
library(stringr)
library(lubridate)

theme_set(
  theme_bw()  
)

get_coda_parameter = function(coda, pattern)
{
  w = coda[[1]] %>% colnames() %>% str_detect(pattern)
  coda[[1]][,w,drop=FALSE]
}

post_summary = function(m, ci_width=0.95)
{
  d = data_frame(
    post_mean  = apply(m, 2, mean),
    post_med   = apply(m, 2, median),
    post_lower = apply(m, 2, quantile, probs=(1-ci_width)/2),
    post_upper = apply(m, 2, quantile, probs=1 - (1-ci_width)/2)
  )
  
  if (!is.null(colnames(m)))
    d = d %>% mutate(param = colnames(m)) %>% dplyr::select(param,post_mean:post_upper)
  
  d
}

strip_attrs = function(obj)
{
  attributes(obj) = NULL
  obj
}

strip_class = function(obj)
{
  attr(obj,"class") = NULL
  obj
}
```


# Spatial Models with AR time dependence

## Example - Weather station data

\footnotesize

Based on Andrew Finley and Sudipto Banerjee's notes from [National Ecological Observatory Network (NEON) Applied Bayesian Regression Workshop, March 7 - 8, 2013](http://blue.for.msu.edu/NEON/SC/) [Module 6](http://blue.for.msu.edu/NEON/SC/exercises/exercise-6/initial-exploration-spDynLM.pdf)

`NETemp.dat` - Monthly temperature data (Celsius) recorded across the Northeastern US starting in January 2000.

\scriptoutput

```{r}
library(spBayes)
data("NETemp.dat")
ne_temp = NETemp.dat %>%
  filter(UTMX > 5.5e6, UTMY > 3e6) %>%
  select(1:27) %>%
  tbl_df()
ne_temp
```

##

```{r echo=FALSE}
ne_temp %>%
  select(1:3,4,7,10,13) %>%
  gather(month, temp, -(elev:UTMY)) %>%
  mutate(month = mdy("1/1/2000") + months(str_extract(month, "[0-9]+") %>% as.integer() - 1)  ) %>%
  ggplot() +
    geom_point(aes(x=UTMX/1000, y=UTMY/1000, color=temp)) +
    facet_wrap(~month)
```

## Dynamic Linear / State Space Models (time)

$$ 
\begin{aligned}
{{y}_t} &= \underset{1 \times p}{\bm{F}'_t} ~ \underset{p \times 1}{\bm{\theta}_t} + {{v}_t} 
&\qquad\qquad\text{observation equation}\\
\underset{p\times 1}{\bm\theta_t} &= \underset{p \times p}{\bm{G}_t} ~ \underset{p \times 1}{\bm\theta}_{t-1} + \underset{p \times 1}{\bm\omega_t}
&\qquad\qquad\text{evolution equation}\\
\end{aligned}
$$

$$ 
\begin{aligned}
\bm{v}_t &\sim \mathcal{N}(0,\bm{V}_t) \\
\bm\omega_t &\sim \mathcal{N}(0,\bm{W}_t) \\
\end{aligned}
$$

## DLM vs ARMA {.t}

ARMA / ARIMA are a special case of a dynamic linear model, for example an $AR(p)$ can be written as
$$ F_t' = (1, 0, \ldots, 0) $$
$$ G_t = \begin{pmatrix}
\phi_1 & \phi_2 & \cdots & \phi_{p-1} & \phi_p \\
1      & 0      & \cdots & 0          & 0      \\
0      & 1      & \cdots & 0          & 0      \\
\vdots & \vdots & \ddots & \vdots     & 0      \\
0      & 0      & \cdots & 1          & 0      \\
\end{pmatrix}$$
$$ \begin{aligned}
\omega_t &= (\omega_1, 0, \ldots 0), 
\quad &\omega_1 \sim \mathcal{N}(0,\,\sigma^2)
\end{aligned}
$$

. . .

\vspace{2mm}

$$
\begin{aligned}
y_t &= \theta_t + v_t,
  \quad &v_t \sim \mathcal{N}(0,\, \sigma^2_v) \\
\theta_t &= \sum_{i=1}^p \phi_i\, \theta_{t-i} + \omega_1, 
  \quad &\omega_1 \sim \mathcal{N}(0,\, \sigma^2_\omega) \\
\end{aligned}
$$


## Dynamic spatio-temporal models

\vspace{2mm}

The observed temperature at time $t$ and location $s$ is given by $y_t(s)$ where,
\footnotesize
$$
\begin{aligned}
y_t(\bm{s}) & = \bm{x}_t(\bm{s})\bm{\beta}_t + u_t(\bm{s}) + \epsilon_t(\bm{s}) \\
\epsilon_t(\bm{s}) &\stackrel{ind.}\sim \mathcal{N}(0,\tau_{t}^2) \\
\\
\bm{\beta}_t & = \bm{\beta}_{t-1} + \bm{\eta}_t \\
\bm{\eta}_t &\stackrel{i.i.d.}\sim \mathcal{N}(0,\bm{\Sigma}_{\eta}) \\
\\
u_t(\bm{s}) &= u_{t-1}(\bm{s}) + w_t(\bm{s}) \\
w_t(\bm{s}) &\stackrel{ind.}{\sim} \mathcal{N}\left(\bm{0}, \Sigma_t(\phi_t, \sigma^2_t)\right)
\end{aligned}
$$

\vspace{3mm}

. . .

\normalsize
Additional assumptions for $t=0$, 
\footnotesize
$$\bm\beta_{0} \sim \mathcal{N}(\bm{\mu}_0, \bm{\Sigma}_0)$$
$$u_{0}(\bm{s}) = 0$$

## Variograms by time

```{r echo=FALSE, message=FALSE}
library(geoR)

coords = ne_temp %>% select(UTMX, UTMY) %>% as.matrix() / 1000

par(mfrow=c(2,2))
variog(coords=coords, data=lm(y.1 ~elev,data=ne_temp)$residuals, uvec=seq(0,300, length.out=30), messages=FALSE) %>% 
  plot(main="Jan 2000", ylim=c(0,5))
variog(coords=coords, data=lm(y.4 ~elev,data=ne_temp)$residuals, uvec=seq(0,300, length.out=30), messages=FALSE) %>% 
  plot(main="Apr 2000", ylim=c(0,5))
variog(coords=coords, data=lm(y.7 ~elev,data=ne_temp)$residuals, uvec=seq(0,300, length.out=30), messages=FALSE) %>% 
  plot(main="Jul 2000", ylim=c(0,5))
variog(coords=coords, data=lm(y.10~elev,data=ne_temp)$residuals, uvec=seq(0,300, length.out=30), messages=FALSE) %>% 
  plot(main="Oct 2000", ylim=c(0,5))
```

## Data and Model Parameters

**Data*:
\scriptoutput
```{r}
coords = ne_temp %>% select(UTMX, UTMY) %>% as.matrix() / 1000
y_t = ne_temp %>% select(starts_with("y.")) %>% as.matrix()

max_d = coords %>% dist() %>% max()
n_t = ncol(y_t)
n_s = nrow(y_t)
```

**Parameters*:
\scriptoutput
```{r}
n_beta = 2
starting = list(
  beta = rep(0, n_t * n_beta), phi = rep(3/(max_d/2), n_t),
  sigma.sq = rep(1, n_t), tau.sq = rep(1, n_t), 
  sigma.eta = diag(0.01, n_beta)
)

tuning = list(phi = rep(1, n_t))

priors = list(
  beta.0.Norm = list(rep(0, n_beta), diag(1000, n_beta)), 
  phi.Unif = list(rep(3/(0.9 * max_d), n_t), rep(3/(0.05 * max_d), n_t)), 
  sigma.sq.IG = list(rep(2, n_t), rep(2, n_t)), 
  tau.sq.IG = list(rep(2, n_t), rep(2, n_t)),
  sigma.eta.IW = list(2, diag(0.001, n_beta))
)
```

## Fitting with `spDynLM` from `spBayes`

\scriptoutput
```{r eval=FALSE}
n_samples = 10000
models = lapply(paste0("y.",1:24, "~elev"), as.formula)

m = spDynLM(
  models, data = ne_temp, coords = coords, get.fitted = TRUE,
  starting = starting, tuning = tuning, priors = priors,
  cov.model = "exponential", n.samples = n_samples, n.report = 1000)

save(m, ne_temp, models, coords, starting, tuning, priors, n_samples, 
     file="dynlm.Rdata")

##  ----------------------------------------
##  	General model description
##  ----------------------------------------
##  Model fit with 34 observations in 24 time steps.
##  
##  Number of missing observations 0.
##  
##  Number of covariates 2 (including intercept if specified).
##  
##  Using the exponential spatial correlation model.
##  
##  Number of MCMC samples 10000.
##
##  ...
```

```{r echo=FALSE}
load(file="dynlm.Rdata")
```

## Posterior Inference - $\beta$s

\vspace{4mm}

```{r echo=FALSE, fig.height=4}
betas = m$p.beta.samples %>% 
  window(start=n_samples/2+1, thin = (n_samples/2)/1000) %>% 
  post_summary() %>%
  mutate(
    month = str_extract(param,"[0-9]+") %>% as.integer(),
    param = str_extract(param,"\\(.*\\)|elev")
  )

ggplot(betas, aes(x=month, y=post_mean, color=param)) +
  geom_point(size=2) +
  geom_linerange(aes(ymin = post_lower, ymax = post_upper)) +
  facet_wrap(~param, scale="free_y")
```

\vvfill

\scriptsize

[Lapse Rate](https://en.wikipedia.org/wiki/Lapse_rate)


## Posterior Inference - $\theta$

\vspace{4mm}

```{r echo=FALSE, fig.height=4}
theta = m$p.theta.samples %>% 
  window(start=n_samples/2+1, thin = (n_samples/2)/1000) %>% 
  post_summary() %>%
  mutate(
    month = str_extract(param,"[0-9]+") %>% as.integer(),
    param = str_extract(param,"sigma\\.sq|tau\\.sq|phi")
  )
eff_range = theta %>% 
  filter(param == "phi") %>%
  mutate(
    param = "Eff. Range",
    post_mean = 3/post_mean,
    post_med = NA,
    tmp = post_lower,
    post_lower = 3/post_upper,
    post_upper = 3/tmp,
    tmp = NULL
  )

theta = rbind(theta, eff_range)

ggplot(theta, aes(x=month, y=post_mean, color=param)) +
  geom_point(size=2) +
  geom_linerange(aes(ymin = post_lower, ymax = post_upper)) +
  facet_wrap(~param, scale="free_y", nrow=3)
```


## Posterior Inference - Observed vs. Predicted

\vspace{4mm}

```{r echo=FALSE, fig.height=5, fig.width=7, fig.align="center",out.width="\\textwidth"}
loc_month = expand.grid(s=1:nrow(ne_temp), month=1:24)

y_hat = m$p.y.samples %>% 
  t() %>% as.mcmc() %>%
  window(start=n_samples/2+1, thin = (n_samples/2)/1000) %>% 
  post_summary() %>%
  cbind(loc_month) %>%
  mutate(y_obs = y_t[loc_month %>% as.matrix()])

ggplot(y_hat, aes(x=y_obs, y=post_mean)) +
  geom_point(size=1, alpha=0.5) +
  ylab("y_hat") +
  #geom_linerange(data=sample_frac(y_hat,0.1), aes(ymin = post_lower, ymax = post_upper), alpha=0.5) +
  geom_abline(slope = 1, intercept=0, color="red", alpha=0.2, size=2) +
  ylim(-20,25) + xlim(-20,25)
```

## Prediction

`spPredict` does not support `spDynLM` objects.

```{r}
r = raster(xmn=575e4, xmx=630e4, ymn=300e4, ymx=355e4, nrow=20, ncol=20)

pred = xyFromCell(r, 1:length(r)) %>% 
  cbind(elev=0, ., matrix(NA, nrow=length(r), ncol=24)) %>%
  as.data.frame() %>%
  setNames(names(ne_temp)) %>%
  rbind(ne_temp, .) %>%
  select(1:15) %>%
  select(-elev)
```  

```{r echo=FALSE}
coords_pred = pred %>% select(UTMX, UTMY) %>% as.matrix()
y_t_pred = pred %>% select(starts_with("y.")) %>% as.matrix()

n_t = ncol(y_t_pred)
n_s = nrow(y_t_pred)

n_beta = 1
starting = list(
  beta = rep(0, n_t * n_beta), phi = rep(3/(max_d/2), n_t),
  sigma.sq = rep(1, n_t), tau.sq = rep(1, n_t), 
  sigma.eta = diag(0.01, n_beta)
)

tuning = list(phi = rep(1, n_t))

priors = list(
  beta.0.Norm = list(rep(0, n_beta), diag(1000, n_beta)), 
  phi.Unif = list(rep(3/(0.9 * max_d), n_t), rep(3/(0.05 * max_d), n_t)), 
  sigma.sq.IG = list(rep(2, n_t), rep(2, n_t)), 
  tau.sq.IG = list(rep(2, n_t), rep(2, n_t)),
  sigma.eta.IW = list(2, diag(0.001, n_beta))
)
```

```{r eval=FALSE}
models_pred = lapply(paste0("y.",1:n_t, "~1"), as.formula)

n_samples = 5000
m_pred = spDynLM(
  models_pred, data = pred, coords = coords_pred, get.fitted = TRUE,
  starting = starting, tuning = tuning, priors = priors,
  cov.model = "exponential", n.samples = n_samples, n.report = 1000)

save(m_pred, pred, models_pred, coords_pred, y_t_pred, n_samples, 
     file="dynlm_pred.Rdata")
```

```{r echo=FALSE}
load("dynlm_pred.Rdata")
```


## 

```{r echo=FALSE}
loc_month = expand.grid(s=1:434, month=1:12)

y_hat_pred = m_pred$p.y.samples %>% 
  t() %>% as.mcmc() %>%
  window(start=n_samples/2+1, thin = 5) %>% 
  post_summary() %>%
  cbind(loc_month) %>%
  mutate(y_obs = y_t_pred[loc_month %>% as.matrix()]) %>%
  mutate(utmx = coords_pred[s,1],
         utmy = coords_pred[s,2])

ggplot(y_hat_pred) +
  geom_point(aes(y=post_mean, x=y_obs)) +
  geom_abline(slope = 1, intercept=0, color="red", alpha=0.2, size=2) +
  ylim(-15,20) + xlim(-15,20)
```

##

```{r echo=FALSE, fig.width=6, fig.align="center", out.width="0.8\\textwidth"}
r_m1 = r
r_m1[] = y_hat_pred %>% filter(s > 34, month==1) %>% .$post_mean

r_m4 = r
r_m4[] = y_hat_pred %>% filter(s > 34, month==4) %>% .$post_mean

r_m7 = r
r_m7[] = y_hat_pred %>% filter(s > 34, month==7) %>% .$post_mean

r_m10 = r
r_m10[] = y_hat_pred %>% filter(s > 34, month==10) %>% .$post_mean


par(mfrow=c(2,2), mar=c(1,1,3,4))
plot(r_m1, main="Jan 2000", axes=FALSE, asp=FALSE); points(coords*1000, pch=16)
plot(r_m4, main="Apr 2000", axes=FALSE, asp=FALSE); points(coords*1000, pch=16)
plot(r_m7, main="Jul 2000", axes=FALSE, asp=FALSE); points(coords*1000, pch=16)
plot(r_m10, main="Oct 2000", axes=FALSE, asp=FALSE); points(coords*1000, pch=16)
```

# Spatio-temporal models for continuous time

## Additive Models {.t}

In general, spatiotemporal models will have a form like the following,

$$ \begin{aligned}
y(\bm{s},{t}) 
  &= \underset{\text{mean structure}}{\mu(\bm{s},{t})} + \underset{\text{error structure}}{{e}(\bm{s},{t})} \\
  &= \underset{\text{Regression}}{\bm{x}(\bm{s},{t}) \, \bm{\beta}(\bm{s},{t})} + \underset{\text{Spatiotemporal RE}}{{w}(\bm{s},{t})} + \underset{\text{Error}}{\epsilon(\bm{s},{t})}
\end{aligned} $$

. . .

\vspace{5mm}

The simplest possible spatiotemporal model is one were assume there is no dependence between observations in space and time,

$$
w(\bm{s},t) = \alpha(t) + \omega(\bm{s})
$$

these are straight forward to fit and interpret but are quite limiting (no shared information between space and time).


## Spatiotemporal Covariance

Lets assume that we want to define our spatiotemporal random effect to be a single stationary Gaussian Process (in 3 dimensions$^\star$),
$$ \bm{w}(\bm{s},\bm{t}) \sim \mathcal{N}\big(\bm{0}, \bm\Sigma(\bm{s},\bm{t})\big) $$
where our covariance function depends on both $\lVert s-s'\rVert$ and $\lvert t-t'\rvert$,
$$
\text{cov}(\bm{w}(\bm{s},\bm{t}), \bm{w}(\bm{s}',\bm{t}')) = c(\lVert s-s'\rVert, \lvert t-t'\rvert)
$$

* Note that the resulting covariance matrix $\Sigma$ will be of size $n_s \cdot n_t \times n_s \cdot n_t$.

    + Even for modest problems this gets very large (past the point of direct computability).

    + If $n_t = 52$ and $n_s = 100$ we have to work with a $5200 \times 5200$ covariance matrix


## Separable Models

One solution is to use a seperable form, where the covariance is the product of a valid 2d spatial and a valid 1d temporal covariance / correlation function,
$$
\text{cov}(\bm{w}(\bm{s},\bm{t}), \bm{w}(\bm{s}',\bm{t}')) = \sigma^2 \, \rho_1(\lVert \bm{s}-\bm{s}'\rVert;\bm\theta) \, \rho_2(\lvert \bm{t}-\bm{t}' \rvert; \bm{\phi})
$$

. . .

If we define our observations as follows (stacking time locations within spatial locations)
\scriptsize
$$
\bm{w}^t_{\bm{s}} = \big(
  w(s_1,t_1)     ,\, \cdots ,\, w(s_1,t_{n_t}) ,\,
  w(s_2,t_1)     ,\, \cdots ,\, w(s_2,t_{n_t}) ,\, \cdots ,\, \cdots ,\,
  w(s_{n_s},t_1) ,\, \cdots ,\, w(s_{n_s},t_{n_t}) \big)
$$
\normalsize
then the covariance can be written as
$$
\bm\Sigma_w(\sigma^2, \theta, \phi) = \sigma^2 \, \bm{H}_s(\theta) \otimes \bm{H}_t(\phi)
$$
where $\bm{H}_s(\theta)$ and $\bm{H}_t(\theta)$ are $n_s \times n_s$ and $n_t \times n_t$ sized correlation matrices respectively and their elements are defined by
\footnotesize
$$\begin{aligned}
\{\bm{H}_s(\theta)\}_{ij} &= \rho_1(\lVert \bm{s}_i - \bm{s}_j \rVert; \theta) \\
\{\bm{H}_t(\phi)\}_{ij} &= \rho_1(\lvert t_i - t_j \rvert; \phi) \\
\end{aligned}$$


## Kronecker Product

Definition:
$$\begin{aligned}
\underset{[m \times n]}{\bm{A}} \otimes \underset{[p \times q]}{\bm{B}} = \underset{[m \cdot p \times  n \cdot q]}{\begin{pmatrix}
a_{11} \bm{B} & \cdots & a_{1n} \bm{B} \\
\vdots        & \ddots & \vdots        \\
a_{m1} \bm{B} & \cdots & a_{mn} \bm{B} \\
\end{pmatrix}}
\end{aligned}$$

\vspace{4mm}

. . .

Properties:
$$\begin{aligned}
\bm{A} \otimes \bm{B}       &\ne \bm{B} \otimes \bm{A}  \qquad\text{(usually)} \\
\vspace{3mm}
(\bm{A} \otimes \bm{B})^t   &= \bm{A}^t \otimes \bm{B}^t \\
\vspace{3mm}
\det(\bm{A} \otimes \bm{B}) &= \det(\bm{B} \otimes \bm{A}) \\
&=\det(\bm{A})^{\text{rank}(\bm{B})} \det(\bm{B})^{\text{rank}(\bm{A})} \\
\vspace{3mm}
(\bm{A} \otimes \bm{B})^{-1} &= \bm{A}^{-1} \bm{B}^{-1}
\end{aligned}$$



## Kronecker Product and MVN Likelihoods

If we have a spatiotemporal random effect with a separable form,
$$
\bm{w}(\bm{s},\bm{t}) \sim \mathcal{N}(\bm{0},\, \bm\Sigma_w)
$$
$$
\bm\Sigma_w = \sigma^2 \, \bm{H}_s \otimes \bm{H}_t
$$

then the likelihood for $\bm{w}$ is given by

$$
-\frac{n}{2}\log 2\pi - \frac{1}{2} \log |\Sigma_w| - \frac{1}{2} \bm{w}^t \bm{\Sigma_w}^{-1} \bm{w}
$$
$$
= -\frac{n}{2}\log 2\pi - \frac{1}{2} \log \left[ (\sigma^2)^{n_t \cdot n_s} |H_s|^{n_t} |H_t|^{n_s}\right] - \frac{1}{2} \bm{w}^t \frac{1}{\sigma^2}(\bm{H}_s^{-1} \otimes \bm{H}_t^{-1}) \bm{w}
$$

## Non-seperable Models {.t}

* Additive and separable models are still somewhat limiting

\vspace{2mm}

* Cannot treat spatiotemporal covariances as 3d observations

\vspace{2mm}

* Possible alternatives:

    + Specialized spatiotemporal covariance functions, i.e.
$$ 
c(\bm{s}-\bm{s}', t-t') 
= \sigma^2 (\lvert t - t'\rvert+1)^{-1} \exp\big(-\lVert\bm{s}-\bm{s}'\rVert (\lvert t-t' \rvert + 1)^{-\beta/2}\big)
$$
$$ ~ $$
    + Mixtures, i.e. $w(\bm{s},t) = w_1(\bm{s},t) + w_2(\bm{s},t)$, where $w_1(\bm{s},t)$ and $w_2(\bm{s},t)$ have seperable forms.
