---
title: "Lecture 4"
author: "Colin Rundel"
date: "1/30/2017"
output: 
  html_document:
    fig_height: 4
---

```{r setup, include=FALSE}
library(rjags)
library(modelr)
library(ggplot2)
library(dplyr)
library(shiny)
library(tidyr)
library(purrr)
library(gridExtra)
```

### Background

Today we'll be looking at data on the presence and absence of the short-finned eel (Anguilla australis) at a number of sites in New Zealand. These data come from Elith, J., Leathwick, J.R., and Hastie, T., 2009. A working guide to boosted regression trees. Journal of Animal Ecology 77: 802-81. 

*Codebook:*

* `presence` - presence or absence of Anguilla australis at the sampling location (0=absent, 1=present)
* `LocSed` -	weighted average of proportional cover of bed sediment (1=mud, 2=sand, 3=fine gravel, 4=coarse gravel, 5=cobble, 6=boulder, 7=bedrock)
* `SegSumT` -	Summer air temperature (degrees C) -->
* `DSDist` - Distance to coast (km)
* `DSDam` -	Presence of known downstream obstructions, mostly dams
* `DSMaxSlope` - Maximum downstream slope (degrees)	
* `USRainDays` - days per month with rain greater than 25 mm
* `USSlope` - average slope in the upstream catchment (degrees)	
* `USNative` - area with indigenous forest (proportion)
* `Method` -	fishing method (`electric`, `net`, `spot`, `trap`, or `mixture`)

```{r}
load(url("http://www.stat.duke.edu/~cr173/Sta444_Sp17/data/anguilla.Rdata"))

set.seed(20170130)

part = resample_partition(anguilla, c(train=0.75, test=0.25))

anguilla = as.data.frame(part$train)
anguilla_test = as.data.frame(part$test)
```


### EDA

```{r warning=FALSE, message=FALSE}
library(GGally)

ggpairs(
  anguilla %>%
    mutate(presence = factor(presence),
           DSDam = factor(DSDam)
    ),
  lower = list(continuous = wrap("points", alpha=0.05))
)

```


### Simple GLM

```{r}
inv_logit = function(x) exp(x)/(1+exp(x))


g = glm(presence~SegSumT, family=binomial, data=anguilla)
#summary(g)

d_g = anguilla %>%
  mutate(p_hat = predict(g, anguilla, type="response"))

d_g_pred = data.frame(SegSumT = seq(11,25,by=0.1)) %>% 
  add_predictions(g,"p_hat") %>%
  mutate(p_hat = inv_logit(p_hat))

ggplot(d_g, aes(x=SegSumT, y=presence)) +
  geom_jitter(height=0.2, alpha=0.5) +
  geom_line(data=d_g_pred, aes(y=p_hat), color="red")
```

#### Residuals

```{r}
d_g = d_g %>%
  mutate(resid = presence - p_hat) %>%
  mutate(p_hat_bin = p_hat - (p_hat %% 0.05))


ggplot(d_g, aes(x=SegSumT, y=resid)) +
  geom_point(alpha=0.3)

d_g %>%
  group_by(p_hat_bin) %>%
  summarize(resid_mean = mean(resid)) %>%
  ggplot(aes(y=resid_mean, x=p_hat_bin)) +
    geom_point()
```

#### Pearson Residuals

$$ r_i = \frac{Y_i - E(Y_i)}{Var(Y_i)} = \frac{Y_i - \hat{p}_i}{\hat{p}_i(1-\hat{p}_i)} $$


```{r}
d_g = d_g %>%
  mutate(pearson = (presence - p_hat) / (p_hat * (1-p_hat)))
         
ggplot(d_g, aes(x=SegSumT, y=pearson)) +
  geom_point(alpha=0.3)   

d_g %>%
  group_by(p_hat_bin) %>%
  summarize(pearson_mean = mean(pearson)) %>%
  ggplot(aes(y=pearson_mean, x=p_hat_bin)) +
    geom_point()
```


#### Deviance Residuals

$$ d_i = \text{sign}(Y_i-\hat{p_i}) \sqrt{ -2 \left(Y_i \log \hat{p}_i+(1-Y_i)\log (1 - \hat{p}_i) \right) }  $$

```{r}

d_g = d_g %>%
  mutate(deviance = sign(presence - p_hat) *sqrt(-2 * (presence*log(p_hat) + (1-presence)*log(1 - p_hat) )))
         
ggplot(d_g, aes(x=SegSumT, y=deviance)) +
  geom_point(alpha=0.3)   

d_g %>%
  group_by(p_hat_bin) %>%
  summarize(deviance_mean = mean(deviance)) %>%
  ggplot(aes(y=deviance_mean, x=p_hat_bin)) +
    geom_point()

```

```{r}
g

sum(d_g$deviance^2)
```


### Full Model

```{r}
f = glm(presence~., family=binomial, data=anguilla)
f
```

#### Residuals vs fitted

```{r}
d_f = anguilla %>%
  mutate(p_hat = predict(f, anguilla, type="response")) %>%
  mutate(
    resid = presence - p_hat,
    pearson = (presence - p_hat) / (p_hat * (1-p_hat)),
    deviance = sign(presence - p_hat) *sqrt(-2 * (presence*log(p_hat) + (1-presence)*log(1 - p_hat) ))
  ) %>%
  mutate(p_hat_bin = p_hat - (p_hat %% 0.05))

d_f_resid = d_f %>%
  select(p_hat:p_hat_bin) %>%
  gather(type, value, -p_hat, -p_hat_bin)
  
ggplot(d_f_resid, aes(x=p_hat, y=value, color=type)) +
  geom_point(alpha=0.3) +
  facet_wrap(~type, ncol=3, scale="free_y")

d_f_resid %>%
  group_by(type, p_hat_bin) %>%
  summarize(
    mean = mean(value)
  ) %>%
  ggplot(aes(x=p_hat_bin, y=mean, color=type)) +
    geom_point() +
    facet_wrap(~type, ncol=3, scale="free_y")
  
```

 
#### Residuals vs predictors

```{r}
ggplot(d_f, aes(x=SegSumT, y=deviance, color=as.factor(presence))) + geom_point()

ggplot(d_f, aes(x=DSMaxSlope, y=deviance, color=as.factor(presence))) + geom_point()
```


### Gradient Boosting model

```{r}
```



### Predictive Performance (ROC / AUC)

```{r}
library(ROCR)

roc = function(d) 
  prediction(d$p_hat, d$presence) %>% performance(measure = "tpr", x.measure = "fpr") 

auc = function(d) 
  {prediction(d$p_hat, d$presence) %>% performance(measure = "auc")}@y.values[[1]]

perf_g  = roc(d_g)
perf_f  = roc(d_f)


plot(perf_g, col="#7fc97f", lwd=2)
plot(perf_f, col="#beaed4",  lwd=2, add=TRUE)
abline(a=0, b = 1, col="grey")


auc(d_g)
auc(d_f)
```



#### Out of sample predictive performance

```{r}

```


