---
title: "Lecture 4"
author: "Colin Rundel"
date: "1/30/2017"
output: 
  html_document:
    fig_height: 4
---

```{r setup, include=FALSE}
library(rjags)
library(modelr)
library(ggplot2)
library(dplyr)
library(shiny)
library(tidyr)
library(purrr)
library(gridExtra)
```

### Background

Today we'll be looking at data on the presence and absence of the short-finned eel (Anguilla australis) at a number of sites in New Zealand. These data come from Elith, J., Leathwick, J.R., and Hastie, T., 2009. A working guide to boosted regression trees. Journal of Animal Ecology 77: 802-81. 

*Codebook:*

* `presence` - presence or absence of Anguilla australis at the sampling location (0=absent, 1=present)
* `LocSed` -	weighted average of proportional cover of bed sediment (1=mud, 2=sand, 3=fine gravel, 4=coarse gravel, 5=cobble, 6=boulder, 7=bedrock)
* `SegSumT` -	Summer air temperature (degrees C) -->
* `DSDist` - Distance to coast (km)
* `DSDam` -	Presence of known downstream obstructions, mostly dams
* `DSMaxSlope` - Maximum downstream slope (degrees)	
* `USRainDays` - days per month with rain greater than 25 mm
* `USSlope` - average slope in the upstream catchment (degrees)	
* `USNative` - area with indigenous forest (proportion)
* `Method` -	fishing method (`electric`, `net`, `spot`, `trap`, or `mixture`)

```{r}
load(url("http://www.stat.duke.edu/~cr173/Sta444_Sp17/data/anguilla.Rdata"))

set.seed(20170130)

part = resample_partition(anguilla, c(train=0.75, test=0.25))

anguilla = as.data.frame(part$train)
anguilla_test = as.data.frame(part$test)
```


### EDA

```{r message=FALSE, echo=FALSE}
library(GGally)

pairs(anguilla)

ggpairs(anguilla)

ggpairs(
  anguilla %>% 
    mutate(
      presence=as.factor(presence),
      DSDam = as.factor(DSDam)
    )
)

ggpairs(
  anguilla %>% 
    mutate(
      presence=as.factor(presence),
      DSDam = as.factor(DSDam)
    ),
  lower = list(continuous = wrap("points", alpha = 0.05))
)
```


### Simple GLM

```{r}
g = glm(presence ~ SegSumT, data=anguilla, family=binomial)
summary(g)
```

```{r}
inv_logit = function(x) exp(x)/(1+exp(x))

pred = data_frame(
  SegSumT = seq(10, 25, length.out = 1000)
) %>%
  add_predictions(g) %>%
  mutate(pred = inv_logit(pred))

ggplot(anguilla, aes(x=SegSumT, y=presence)) +
  #geom_point() +
  geom_jitter(height=0.05, alpha=0.33) + 
  geom_line(data=pred, aes(y=pred), col='red')
```

#### Residuals

```{r}
d_g = anguilla %>% 
  add_predictions(g, var="p_hat") %>%
  mutate(p_hat = inv_logit(p_hat)) %>%
  mutate(resid = presence - p_hat) %>%
  mutate(p_hat_bin = p_hat - (p_hat %% 0.025))

ggplot(d_g, aes(x=p_hat, y=resid)) +
    geom_point()

d_g %>% 
  group_by(p_hat_bin) %>%
  summarize(resid_mean = mean(resid), resid_sd = sd(resid)) %>%
  ggplot(aes(x=p_hat_bin, y=resid_mean)) +
    geom_point()
```

#### Pearson Residuals

$$ r_i = \frac{Y_i - E(Y_i)}{Var(Y_i)} = \frac{Y_i - \hat{p}_i}{\hat{p}_i(1-\hat{p}_i)} $$

```{r}
d_g = d_g %>%
  mutate(pearson = (presence - p_hat)/(p_hat*(1-p_hat)))

ggplot(d_g, aes(x=p_hat, y=pearson)) +
  geom_point()

d_g %>% 
  group_by(p_hat_bin) %>%
  summarize(pearson_mean = mean(pearson), resid_sd = sd(pearson)) %>%
  ggplot(aes(x=p_hat_bin, y=pearson_mean)) +
    geom_point()
```


#### Deviance Residuals

$$ d_i = \text{sign}(Y_i-\hat{p_i}) \sqrt{ -2 \left(Y_i \log \hat{p}_i+(1-Y_i)\log (Y_i - \hat{p}_i) \right) }  $$

```{r}
d_g = d_g %>%
  mutate(deviance = sign(presence-p_hat) * sqrt(-2 * (presence * log(p_hat) + (1-presence) * log(1-p_hat))))

ggplot(d_g, aes(x=p_hat, y=deviance)) +
  geom_point()

d_g %>% 
  group_by(p_hat_bin) %>%
  summarize(deviance_mean = mean(deviance), deviance_sd = sd(deviance)) %>%
  ggplot(aes(x=p_hat_bin, y=deviance_mean)) +
    geom_point()
```

```{r}
g
sum(d_g$deviance^2)
```

#### Shiny residuals

```{r eval=FALSE}
shinyApp(
  ui = mainPanel(
    plotOutput("plot"),
    selectInput("var", label = "Variable", choices = c("resid","deviance","pearson")),
    sliderInput("bin_width", label="Bin width", min = 0, max=0.2, step = 0.005, value=0.025)
  ),
  server = function(input, output)
  {
    output$plot = renderPlot(
      d_g %>% 
        mutate(p_hat_bin = p_hat - (p_hat %% input$bin_width)) %>%
        group_by(p_hat_bin) %>%
        summarize_(mean = paste0('mean(', input$var, ')')) %>%
        ggplot(aes(x=p_hat_bin, y=mean)) +
          geom_point()
    )
  }
)
```


### Full Model

```{r}
f = glm(presence~., data=anguilla, family=binomial)
summary(f)
```

#### Residuals vs fitted

```{r}
d_f = anguilla %>%
  add_predictions(f, var="p_hat") %>%
  mutate(p_hat = inv_logit(p_hat)) %>%
  mutate(p_hat_bin = p_hat - (p_hat %% 0.025)) %>%
  mutate(
    resid = presence - p_hat,
    pearson = (presence - p_hat)/(p_hat*(1-p_hat)),
    deviance = sign(presence-p_hat) * sqrt(-2 * (presence * log(p_hat) + (1-presence) * log(1-p_hat)))
  )

resid_f = d_f %>%
  select(p_hat:deviance) %>%
  gather(type, value, -p_hat, -p_hat_bin) %>%
  mutate(type = factor(type, levels=c("resid","pearson","deviance"))) 

ggplot(resid_f, aes(x=p_hat, y=value, color=type)) +
  geom_point(alpha=0.1) +
  facet_wrap(~type, ncol=3, scale="free_y")
```

 
```{r}
resid_f %>%
  group_by(type, p_hat_bin) %>%
  summarize(mean = mean(value)) %>%
  ggplot(aes(x=p_hat_bin, y=mean, color=type)) +
    geom_point() +
    facet_wrap(~type, ncol=3, scale="free_y")
```

#### Residuals vs predictors

```{r}
ggplot(d_f, aes(x=SegSumT, y=deviance, color=as.factor(presence))) + geom_point()
ggplot(d_f, aes(x=DSMaxSlope, y=deviance, color=as.factor(presence))) + geom_point()
```

Possible fix:

```{r}
f2 = glm(presence~.+I(SegSumT^2), data=anguilla, family=binomial)
summary(f2)

d_f2 = anguilla %>%
  add_predictions(f2, var="p_hat") %>%
  mutate(p_hat = inv_logit(p_hat)) %>%
  mutate(p_hat_bin = p_hat - (p_hat %% 0.025)) %>%
  mutate(
    resid = presence - p_hat,
    pearson = (presence - p_hat)/(p_hat*(1-p_hat)),
    deviance = sign(presence-p_hat) * sqrt(-2 * (presence * log(p_hat) + (1-presence) * log(1-p_hat)))
  )

rbind(
  cbind(d_f,  model="linearT"),
  cbind(d_f2, model="quadT")
) %>%
  ggplot(aes(x=SegSumT, y=deviance, color=as.factor(presence))) + geom_point() + facet_wrap(~model)
```


```{r message=FALSE}
d_f %>% 
  select(deviance, SegSumT:LocSed) %>%
  mutate(DSDam = factor(DSDam)) %>%
  ggpairs(
    lower = list(continuous = "cor"),
    upper = list(continuous = wrap("points", alpha = 0.05))
  )
```

### Gradient Boosting

```{r}
library(xgboost)

y = anguilla$presence %>% as.integer()
x = model.matrix(presence~.-1, data=anguilla)

xg = xgboost(data=x, label=y, nthead=4, nround=30, objective="binary:logistic")

pred_xg = 

d_xg = anguilla %>%
  mutate(p_hat = predict(xg, newdata=x)) %>%
  mutate(p_hat_bin = p_hat - (p_hat %% 0.025)) %>%
  mutate(
    resid = presence - p_hat,
    pearson = (presence - p_hat)/(p_hat*(1-p_hat)),
    deviance = sign(presence-p_hat) * sqrt(-2 * (presence * log(p_hat) + (1-presence) * log(1-p_hat)))
  )

resid_xg = d_xg %>%
  select(p_hat:deviance) %>%
  gather(type, value, -p_hat, -p_hat_bin) %>%
  mutate(type = factor(type, levels=c("resid","pearson","deviance"))) 

ggplot(resid_xg, aes(x=p_hat, y=value, color=type)) +
  geom_point(alpha=0.1) +
  facet_wrap(~type, ncol=3, scale="free_y")

resid_f %>%
  group_by(type, p_hat_bin) %>%
  summarize(mean = mean(value)) %>%
  ggplot(aes(x=p_hat_bin, y=mean, color=type)) +
    geom_point() +
    facet_wrap(~type, ncol=3, scale="free_y")
```



### Predictive Performance

```{r}
grid.arrange(
  ggplot(d_g, aes(x=p_hat, y=factor(presence), color=factor(presence))) +
    geom_jitter(height=0.2, alpha=0.5) + labs(title="simple"),
  
  ggplot(d_f, aes(x=p_hat, y=factor(presence), color=factor(presence))) +
    geom_jitter(height=0.2, alpha=0.5) + labs(title="full"),
  
  ggplot(d_xg, aes(x=p_hat, y=factor(presence), color=factor(presence))) +
    geom_jitter(height=0.2, alpha=0.5) + labs(title="xgboost")
)
```

```{r}
score = function(p_hat, label, cutoff)
{
  map_df(
    cutoff, 
    ~ data.frame(
      sensitivity = sum(p_hat >= . &  label) / length(p_hat),
      specificity = sum(p_hat <  . & !label) / length(p_hat)
    ) 
  )
}

score(d_f$p_hat, d_f$presence, seq(0,1,by=0.1)) %>%
  ggplot(aes(x=specificity, y=sensitivity)) +
  geom_point() + 
  geom_line()

score(d_f$p_hat, d_f$presence, seq(0,1,by=0.01)) %>%
  ggplot(aes(x=1-specificity, y=sensitivity)) +
  #geom_point() + 
  geom_line()
```

```{r}
library(ROCR)

roc = function(d) 
  prediction(d$p_hat, d$presence) %>% performance(measure = "tpr", x.measure = "fpr") 

auc = function(d) 
  {prediction(d$p_hat, d$presence) %>% performance(measure = "auc")}@y.values[[1]]

perf_g  = roc(d_g)
perf_f  = roc(d_f)
perf_xg = roc(d_xg)

plot(perf_g, col="#7fc97f", lwd=2)
plot(perf_f, col="#beaed4",  lwd=2, add=TRUE)
plot(perf_xg, col="#fdc086",  lwd=2, add=TRUE)
abline(a=0, b = 1, col="grey")


auc(d_g)
auc(d_f)
auc(d_xg)
```

#### Out of sample ROC

```{r}

d_f_test = anguilla_test %>%
  add_predictions(f, var="p_hat") %>%
  mutate(p_hat = inv_logit(p_hat)) %>%
  mutate(p_hat_bin = p_hat - (p_hat %% 0.025)) %>%
  mutate(
    resid = presence - p_hat,
    pearson = (presence - p_hat)/(p_hat*(1-p_hat)),
    deviance = sign(presence-p_hat) * sqrt(-2 * (presence * log(p_hat) + (1-presence) * log(1-p_hat)))
  )


d_xg_test = anguilla_test %>%
  mutate(p_hat = predict(xg, newdata=model.matrix(presence~.-1, data=anguilla_test))) %>%
  mutate(p_hat_bin = p_hat - (p_hat %% 0.025)) %>%
  mutate(
    resid = presence - p_hat,
    pearson = (presence - p_hat)/(p_hat*(1-p_hat)),
    deviance = sign(presence-p_hat) * sqrt(-2 * (presence * log(p_hat) + (1-presence) * log(1-p_hat)))
  )
```


```{r}

grid.arrange(
  ggplot(d_f_test, aes(x=p_hat, y=factor(presence), color=factor(presence))) +
    geom_jitter(height=0.2, alpha=0.5) + labs(title="full"),
  
  ggplot(d_xg_test, aes(x=p_hat, y=factor(presence), color=factor(presence))) +
    geom_jitter(height=0.2, alpha=0.5) + labs(title="xgboost")
)
```

```{r}
plot(roc(d_f_test), col="#beaed4",  lwd=2)
plot(roc(d_xg_test), col="#fdc086",  lwd=2, add=TRUE)
abline(a=0, b = 1, col="grey")

auc(d_f_test)
auc(d_xg_test)
```


