
---
title: "Lecture 6" 
subtitle: "Discrete Time Series"
author: "Colin Rundel"
date: "02/06/2017"
fontsize: 11pt
output: 
  beamer_presentation:
    theme: metropolis
    highlight: pygments
    fig_width: 8
    fig_height: 6
    fig_caption: false
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: ../settings.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)

set.seed(20170206)

library(magrittr)
library(dplyr)
library(modelr)
library(ggplot2)
library(tidyr)
library(rjags)
library(stringr)
library(gridExtra)
library(readr)

theme_set(
  theme_bw()  
  #theme(plot.title = element_text(hjust = 0.5))
)

get_coda_parameter = function(coda, pattern)
{
  w = coda[[1]] %>% colnames() %>% str_detect(pattern)
  coda[[1]][,w,drop=FALSE]
}

post_summary = function(m, ci_width=0.95)
{
  data_frame(
    post_mean  = apply(m, 2, mean),
    post_lower = apply(m, 2, quantile, probs=(1-ci_width)/2),
    post_upper = apply(m, 2, quantile, probs=1 - (1-ci_width)/2)
  )
}

```

# Discrete Time Series

## Stationary Processes {.t}

A stocastic process (i.e. a time series) is considered to be *strictly stationary* if the properties of the process are not changed by a shift in origin. 

. . .

In the time series context this means that the joint distribution of $\{y_{t_1}, \ldots, y_{t_n}\}$ must be identical  to the distribution of $\{y_{t_1+k}, \ldots, y_{t_n+k}\}$ for any value of $n$ and $k$.

. . . 



## Weak Stationary {.t}

Strict stationary is too strong for most applications, so instead we often opt for *weak stationary* which requires the following,

1. The process has finite variance 
$$E(y_t^2) < \infty \text{ for all $t$}$$

2. The mean of the process in constant 
$$E(y_t) = \mu \text{ for all $t$}$$

3. The second moment only depends on the lag 
$$Cov(y_t,y_s) = Cov(y_{t+k},y_{s+k}) \text{ for all $t,s,k$}$$

. . . 

When we say stationary in class we almost always mean this version of *weakly stationary*.

## Autocorrelation {.t}

For a stationary time series, where $E(y_t)=\mu$ and $\text{Var}(y_t)=\sigma^2)$ for all $t$, we define the autocorrelation at lag $k$ as

$$
\begin{aligned}
\rho_k &= Cor(y_t, \, y_{t+k}) \\
       &= \frac{Cov(y_t, y_{t+k})}{\sqrt{Var(y_t)Var(y_{t+k})}} \\
       &= \frac{E\left( (y_t-\mu)(y_{t+k}-\mu) \right)}{\sigma^2}
\end{aligned}
$$

. . .

this is also sometimes written in terms of the autocovariance function ($\gamma_k$) as

$$
\begin{aligned}
\gamma_k &= \gamma(t,t+k) = Cov(y_t, y_{t+k}) \\
\rho_k &= \frac{\gamma(t,t+k)}{\sqrt{\gamma(t,t) \gamma(t+k,t+k)}} = \frac{\gamma(k)}{\gamma(0)}
\end{aligned}
$$

## Covariance Structure

Based on our definition of a (weakly) stationary process, it implies a covariance of the following structure,

$$
\left(
\begin{matrix}
\gamma(0) & \gamma(1) & \gamma(2) & \gamma(3) & \cdots & \gamma(n) \\
\gamma(1) & \gamma(0) & \gamma(1) & \gamma(2) & \cdots & \gamma(n-1) \\
\gamma(2) & \gamma(1) & \gamma(0) & \gamma(1) & \cdots & \gamma(n-2) \\
\gamma(3) & \gamma(2) & \gamma(1) & \gamma(0) & \cdots & \gamma(n-3) \\
\vdots    & \vdots    & \vdots    & \vdots    & \ddots & \vdots \\
\gamma(n) & \gamma(n-1) & \gamma(n-2) & \gamma(n-3) & \cdots & \gamma(0)
\end{matrix}
\right)
$$


## Example - Random walk {.t}

Let $y_t = y_{t-1} + w_t$ with $y_0=0$ and $w_t \sim \mathcal{N}(0,1)$. Is $y_t$ stationary?

<!--
$$
y_0 = 0 \\
y_1 = \epsilon_1 \\
y_0 = \epsilon_1 + \epsilon_2 \\
y_0 =  \epsilon_1 + \epsilon_2 + \epsilon_3 \\
\cdots \\
y_t = \sum_{t=1}^n \epsilon_t \\
$$

$$ E(y_t) = \sum 0 = 0 $$

$$ Cov(y_t, y_{t+k}) = E(y_t \times y_{t+k}) = t $$
-->

```{r echo=FALSE}
rw = data_frame(
  t = 1:1000,
  w = c(0, rnorm(999))
) %>%
  mutate(
    y = cumsum(w)
  )

ggplot(rw, aes(x=t, y=y)) + geom_line() + labs(title="Random walk")
```

## ACF + PACF

```{r echo=FALSE, fig.height=4}
par(mfrow=c(1,2))
acf(rw$y, lag.max = 50)
pacf(rw$y, lag.max = 50)
```


## Example - Random walk with drift {.t}

Let $y_t = \delta + y_{t-1} + w_t$ with $y_0=0$ and $w_t \sim \mathcal{N}(0,1)$. Is $y_t$ stationary?

```{r echo=FALSE}
rwt = data_frame(
  t = 1:1000,
  w = c(0, 0.1+rnorm(999))
) %>%
  mutate(
    y = cumsum(w)
  )

ggplot(rwt, aes(x=t, y=y)) + geom_line() + labs(title="Random walk with trend")
```

## ACF + PACF

```{r echo=FALSE, fig.height=4}
par(mfrow=c(1,2))
acf(rwt$y, lag.max = 50)
pacf(rwt$y, lag.max = 50)
```

## Example - Moving Average

Let $w_t \sim \mathcal{N}(0,1)$ and $y_t = \frac{1}{3}\left(w_{t-1}+w_t+w_{t+1}\right)$, is $y_t$ stationary?

```{r echo=FALSE, warning=FALSE}
ma = data_frame(
  t = 1:100,
  w = rnorm(100)
) %>%
  mutate(
    y = (c(NA,w[-100]) + w + c(w[-1],NA))/3
  )

ggplot(ma, aes(x=t, y=y)) + geom_line() + labs(title="Moving Average")
```

## ACF + PACF

```{r echo=FALSE, fig.height=4}
par(mfrow=c(1,2))
acf(ma$y, lag.max = 50, na.action = na.omit)
pacf(ma$y, lag.max = 50, na.action = na.omit)
```

## Autoregression

Let $w_t \sim \mathcal{N}(0,1)$ and $y_t = y_{t-1} - 0.9 y_{t-2} + w_t$ with $y_t = 0$ for $t < 1$, is $y_t$ stationary?

```{r echo=FALSE, warning=FALSE}
ar = data_frame(
  t = 1:500,
  w = rnorm(500),
  y = NA
)

for(i in seq_along(ar$w))
{
  if (i == 1)
    ar$y[i] = ar$w[i]
  else if (i==2)
    ar$y[i] = ar$y[i-1] + ar$w[i]
  else
    ar$y[i] = ar$y[i-1] -0.9*ar$y[i-2] + ar$w[i]
}
  
ggplot(ar, aes(x=t, y=y)) + geom_line() + labs(title="Autoregressive")
```


## ACF + PACF

```{r echo=FALSE, fig.height=4}
par(mfrow=c(1,2))
acf(ar$y, lag.max = 50, na.action = na.omit)
pacf(ar$y, lag.max = 50, na.action = na.omit)
```

## Example - Australian Wine Sales

Australian total wine sales by wine makers in bottles <= 1 litre. Jan 1980 – Aug 1994.

```{r}
load(url("http://www.stat.duke.edu/~cr173/Sta444_Sp17/data/aus_wine.Rdata"))
aus_wine
```

## Time series

```{r echo=FALSE}
ggplot(aus_wine, aes(x=date, y=sales)) + 
  geom_line() + 
  geom_point()
```

## Basic Model Fit

```{r echo=FALSE}
l  = lm(sales ~ date, data=aus_wine)
l2 = lm(sales ~ date + I(date^2), data=aus_wine)

d = aus_wine %>%
  add_predictions(l, var="l") %>%
  add_predictions(l2, var="l2")

ggplot(aus_wine, aes(x=date, y=sales)) + 
  geom_line() + 
  geom_point() +
  geom_line(data=d, aes(x=date, y=l), color="red", size=1.2) +
  geom_line(data=d, aes(x=date, y=l2), color="blue", size=1.2)
```

## Residuals

```{r echo=FALSE}
d = aus_wine %>%
  add_residuals(l, var="resid_l") %>%
  add_residuals(l2, var="resid_q")
  

ggplot(gather(d, type, residual, -(date:sales)), aes(x=date, y=residual, color=type)) + 
  geom_point() +
  geom_line() +
  facet_wrap(~type, nrow=2)
```




## Autocorrelation Plot

```{r echo=FALSE}
acf(d$resid_q, lag.max = 36)
```

## Partial Autocorrelation Plot

```{r echo=FALSE}
pacf(d$resid_q, lag.max = 36)
```

##

```{r echo=FALSE, message=FALSE, warning=FALSE}
d_lags = d %>%
  mutate(
    lag_01 = lag(resid_q, 1),
    lag_02 = lag(resid_q, 2),
    lag_03 = lag(resid_q, 3),
    lag_04 = lag(resid_q, 4),
    lag_05 = lag(resid_q, 5),
    lag_06 = lag(resid_q, 6),
    lag_07 = lag(resid_q, 7),
    lag_08 = lag(resid_q, 8),
    lag_09 = lag(resid_q, 9),
    lag_10 = lag(resid_q, 10),
    lag_11 = lag(resid_q, 11),
    lag_12 = lag(resid_q, 12)
  ) %>% gather(lag, lag_value, -(date:resid_q))

ggplot(d_lags, aes(x=lag_value, y=resid_q)) +
  geom_point() +
  facet_wrap(~lag, ncol=4) +
  geom_smooth(method="lm", color='red', se = FALSE, alpha=0.1)
```




## Auto regressive errors

```{r echo=FALSE}
d_ar = mutate(d, lag_12 = lag(resid_q, 12))
l_ar = lm(resid_q ~ lag_12, data=d_ar)
summary(l_ar)
```

## Residual residuals

```{r echo=FALSE, warning=FALSE}
d_ar %>%
  add_residuals(l_ar) %>%
  ggplot(aes(x=date, y=resid)) +
  geom_point() + 
  geom_line()
```

## Residual residuals - acf

```{r echo=FALSE}
acf(l_ar$residuals, lag.max = 36)
```

##

```{r echo=FALSE, warning=FALSE, fig.height=5.5}
d_ar %>%
  add_residuals(l_ar) %>%
  select(-lag_12) %>%
  mutate(
    lag_01 = lag(resid, 1),
    lag_02 = lag(resid, 2),
    lag_03 = lag(resid, 3),
    lag_04 = lag(resid, 4),
    lag_05 = lag(resid, 5),
    lag_06 = lag(resid, 6),
    lag_07 = lag(resid, 7),
    lag_08 = lag(resid, 8),
    lag_09 = lag(resid, 9),
    lag_10 = lag(resid, 10),
    lag_11 = lag(resid, 11),
    lag_12 = lag(resid, 12)
  ) %>% 
  gather(lag, lag_value, -(date:resid)) %>%
  ggplot(aes(x=lag_value, y=resid)) +
  geom_point() +
  facet_wrap(~lag, ncol=4) +
  geom_smooth(method="lm", color='red', se = FALSE, alpha=0.1)
```

## Writing down the model? {.t}

So, is our EDA suggesting that we then fit the following model?

$$ \text{sales}(t) = \beta_0 + \beta_1 \, t + \beta_2 \, t^2 + \beta_3 \, \text{sales}(t-12) + \epsilon_t $$
. . .

the implied model is,

$$ \text{sales}(t) = \beta_0 + \beta_1 \, t + \beta_2 \, t^2 + w_t $$
where

$$ w_t = \delta \, w_{t-12} + \epsilon_t $$






