
---
title: "Lecture 17" 
subtitle: "Spatial Data and Cartography (Part 2)"
author: "Colin Rundel"
date: "03/22/2017"
fontsize: 11pt
output: 
  beamer_presentation:
    theme: metropolis
    highlight: pygments
    fig_width: 8
    fig_height: 5
    fig_caption: false
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: ../settings.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning=FALSE, message=FALSE, 
                      fig.align="center", out.width="0.95\\textwidth")
options(width=70)

set.seed(20170322)

library(raster)
library(dplyr)
library(sf)
library(stringr)
library(purrr)
library(ggplot2)
library(gridExtra)


strip_attrs = function(obj)
{
  attributes(obj) = NULL
  obj
}

strip_class = function(obj)
{
  attr(obj,"class") = NULL
  obj
}
```

# Plotting

## Example Data - NC SIDS {.t}

\scriptoutput

```{r}
nc = st_read(system.file("shape/nc.shp", package="sf"), quiet = TRUE) %>% 
  select(-(AREA:CNTY_ID), -(FIPS:CRESS_ID))

head(nc, n = 10)
```

## Base Plots {.t}

\scriptoutput

```{r out.width="0.8\\textwidth"}
plot(nc)
```

## Geometry Plot {.t}

\scriptoutput

```{r}
plot(st_geometry(nc), axes=TRUE)
```

## Graticules {.t}

\scriptoutput

```{r}
plot(nc[,"SID79"], graticule=st_crs(nc), axes=TRUE, las=1)
```

## Graticules (EPSG:3631) {.t}

\scriptoutput

```{r}
plot(st_transform(nc[,"SID79"], 3631), graticule=st_crs(nc), axes=TRUE, las=1)
```


## ggplot2 2.2.1.9 (dev) {.t}

\scriptoutput

```{r}
ggplot(nc) + 
  geom_sf(aes(fill=SID79 / BIR79))
```

## ggplot2 + projections {.t}

\scriptoutput

```{r}
ggplot(st_transform(nc, 3631)) + 
  geom_sf(aes(fill=SID79 / BIR79))
```


## Example Data - Meuse

\scriptoutput

```{r}
data(meuse, meuse.riv, package="sp")

meuse = st_as_sf(meuse, coords=c("x", "y"), crs=28992)
meuse_riv = st_polygon(list(meuse.riv)) %>% st_sfc() %>% st_set_crs(28992)

meuse
```

## Meuse {.t}

\scriptoutput

```{r}
plot(meuse, pch=16)
```



## Layering plots {.t}

\scriptoutput

```{r}
plot(meuse[,"lead"], pch=16, axes=TRUE)
plot(meuse_riv, col=adjustcolor("lightblue", alpha.f=0.5), add=TRUE, border = NA)
```

## Layering plots (oops)  {.t}

\scriptoutput

```{r}
plot(meuse, pch=16)
plot(meuse_riv, col=adjustcolor("lightblue", alpha.f=0.5), add=TRUE, border = NA)
```

## ggplot2 {.t}

\scriptoutput

```{r out.width="0.33\\textwidth"}
ggplot() +
  geom_sf(data=st_sf(meuse_riv), fill="lightblue", color=NA) +
  geom_sf(data=meuse, aes(color=lead), size=1)
```

## ggplot2 - axis limits {.t}

\scriptoutput

```{r out.width="0.65\\textwidth"}
ggplot() +
  geom_sf(data=st_sf(meuse_riv), fill="lightblue", color=NA) +
  geom_sf(data=meuse, aes(color=lead), size=0.1) +
  ylim(329714, 333611)
```

## ggplot2 - bounding box{.t}

\scriptoutput

```{r out.width="0.65\\textwidth"}
ggplot() +
  geom_sf(data=st_sf(meuse_riv), fill="lightblue", color=NA) +
  geom_sf(data=meuse, aes(color=lead), size=0.1) +
  ylim(st_bbox(meuse)["ymin"], st_bbox(meuse)["ymax"])
```

# Geometry Manipulation

## Casting {.t}

\scriptoutput

```{r}
nc_pts = st_cast(nc, "MULTIPOINT")

nc_pts
```

## {.t}

\scriptoutput

```{r}
plot(st_geometry(nc), border='grey')
plot(st_geometry(nc_pts), pch=16, cex=0.5, add=TRUE)
```

## Casting - POINT {.t}

\scriptoutput

```{r}
st_cast(nc, "POINT")
```

## Casting - LINESTRING {.t}

\scriptoutput

```{r}
st_cast(nc, "LINESTRING")
```

## {.t}

\scriptoutput

```{r}
st_cast(nc, "LINESTRING") %>% st_geometry() %>% plot()
```


## Grouping Features {.t}

\scriptoutput

```{r}
nc_state = st_union(nc)
plot(nc_state)

nc_state
```

## More Grouping {.t}

\scriptoutput

```{r}
nc_cut = nc %>%
  cbind(nc %>% st_centroid() %>% st_coordinates()) %>%
  mutate(region = cut(X, breaks = 5))

nc_cut
```

## {.t}

\scriptoutput

```{r}
ggplot(nc_cut) +
  geom_sf(aes(fill=region))
```

## dplyr and sf - BFFs {.t}

\scriptoutput

```{r}
nc_cut %>% 
  group_by(region) %>% 
  summarize(geometry = st_union(geometry)) %>% 
  ggplot() + geom_sf(aes(fill=region))
```


## Affine Transfomations {.t}

\scriptoutput

```{r}
rotate = function(a) matrix(c(cos(a), sin(a), -sin(a), cos(a)), 2, 2)

ctrd = st_centroid(nc_state)
state_rotate = (nc_state - ctrd) * rotate(-pi/4) + ctrd
plot(state_rotate, axes=TRUE)
```

## Scaling Size

\scriptoutput

```{r}
ctrd = st_geometry(st_centroid(nc))
area = st_area(nc) %>% strip_attrs()
nc_scaled = nc
for(i in 1:nrow(nc))
  nc_scaled$geometry[[i]] = ((st_geometry(nc[i,]) - ctrd[i]) *
                               sqrt(min(area)/area[i]) + ctrd[i])[[1]]

plot(nc_scaled[,"SID79"])
```


## Back to the highways {.t}

\scriptoutput

```{r}
hwy = st_read("../../data/gis/us_interstates/", quiet=TRUE, stringsAsFactors=FALSE) %>% st_transform(st_crs(nc))

ggplot() +
  geom_sf(data=nc) +
  geom_sf(data=hwy, col='red')
```

## NC Interstate Highways {.t}

\scriptoutput

```{r}
hwy_nc = st_intersection(hwy, nc)

ggplot() + 
  geom_sf(data=nc) +
  geom_sf(data=hwy_nc, col='red')

hwy_nc
```

## Counties near the interstate (Projection) {.t}

\scriptoutput

```{r}
nc_utm  = st_transform(nc,  "+proj=utm +zone=17 +datum=NAD83 +units=m +no_defs")
hwy_utm = st_transform(hwy, "+proj=utm +zone=17 +datum=NAD83 +units=m +no_defs")

hwy_nc_utm = st_intersection(nc_utm, hwy_utm)

ggplot() + 
  geom_sf(data=nc_utm) +
  geom_sf(data=hwy_nc_utm, col='red')
```

## Counties near the interstate (Buffering) {.t}

\scriptoutput

```{r}
hwy_nc_buffer = st_buffer(hwy_nc_utm, 10000)

ggplot() + 
  geom_sf(data=nc_utm) +
  geom_sf(data=hwy_nc_utm, color='red') +
  geom_sf(data=hwy_nc_buffer, fill='red', alpha=0.3)
```

## Counties near the interstate (Buffering + Union) {.t}

\scriptoutput

```{r}
hwy_nc_buffer = st_buffer(hwy_nc_utm, 10000) %>% st_union() %>% st_sf()

ggplot() + 
  geom_sf(data=nc_utm) +
  geom_sf(data=hwy_nc_utm, color='red') +
  geom_sf(data=hwy_nc_buffer, fill='red', alpha=0.3)
```

## Exercise 1 {.t}

How many counties in North Carolina are within 5, 10, 20, or 50 km of an interstate highway?



# Raster Data

## Example data - Meuse {.t}

\scriptoutput

```{r}
meuse_rast = raster(system.file("external/test.grd", package="raster"))

meuse_rast
```

## {.t}

\scriptoutput

```{r}
plot(meuse_rast)
plot(meuse_riv, add=TRUE, col=adjustcolor("lightblue",alpha.f = 0.5), border=NA)
```

## raster class

```{r}
str(meuse_rast)
```

## raster features {.t}

\scriptoutput

```{r}
extent(meuse_rast)

dim(meuse_rast)

res(meuse_rast)

projection(meuse_rast)

meuse_rast[20,]
```

## Rasters and Projections {.t}

\scriptoutput

```{r}
meuse_rast_ll = projectRaster(meuse_rast, crs="+proj=longlat +datum=NAD27 +no_defs")

par(mfrow=c(1,2))
plot(meuse_rast)
plot(meuse_rast_ll)
```

## {.t}

\scriptoutput

```{r}
meuse_rast

meuse_rast_ll
```



## Simple Features $\longleftrightarrow$ Rasters {.t}

\scriptoutput

```{r error=TRUE}
meuse_riv_rast = rasterize(meuse_riv, meuse_rast)

meuse_riv_rast = rasterize(as(meuse_riv, "Spatial"), meuse_rast)
plot(meuse_riv_rast)
```


## {.t}

\scriptoutput

```{r error=TRUE}
sub = !is.na(meuse_riv_rast[]) & !is.na(meuse_rast[])
river_obs = meuse_rast
river_obs[!sub] = NA
plot(river_obs)

xyFromCell(river_obs, which(sub))
```

## Rasters and Spatial Models {.t}

\scriptoutput

```{r}
head(meuse)

head(st_coordinates(meuse))
```

## {.t}

\scriptoutput

```{r}
library(fields)

tps = Tps(x = st_coordinates(meuse), Y=meuse$elev)
pred_grid = xyFromCell(meuse_rast, seq_along(meuse_rast))

meuse_elev_pred = meuse_rast
meuse_elev_pred[] = predict(tps, pred_grid)

plot(meuse_elev_pred)
```

## Hacky Crap

```{r fig.height=4}
p = rasterToPolygons(meuse_elev_pred) %>% st_as_sf()
grid.arrange(
  ggplot() + geom_sf(data=meuse, aes(color=elev), size=0.1),
  ggplot() + geom_sf(data=p, aes(fill=test), color=NA),
  ncol=2
)
```
