
---
title: "Lecture 15" 
subtitle: "Mauna Loa Example & GPs for GLMs"
author: "Colin Rundel"
date: "03/08/2017"
fontsize: 11pt
output: 
  beamer_presentation:
    theme: metropolis
    highlight: pygments
    fig_width: 8
    fig_height: 5
    fig_caption: false
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: ../settings.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning=FALSE, message=FALSE)
options(width=70)

set.seed(20170308)

library(magrittr)
library(dplyr)
library(modelr)
library(ggplot2)
library(tidyr)
library(rjags)
library(stringr)
library(gridExtra)
library(readr)
library(purrr)
library(forcats)
library(forecast)
library(astsa)
library(fields)
library(readr)

theme_set(
  theme_bw()  
)

get_coda_parameter = function(coda, pattern)
{
  w = coda[[1]] %>% colnames() %>% str_detect(pattern)
  coda[[1]][,w,drop=FALSE]
}

post_summary = function(m, ci_width=0.95)
{
  d = data_frame(
    post_mean  = apply(m, 2, mean),
    post_med   = apply(m, 2, median),
    post_lower = apply(m, 2, quantile, probs=(1-ci_width)/2),
    post_upper = apply(m, 2, quantile, probs=1 - (1-ci_width)/2)
  )
  
  if (!is.null(colnames(m)))
    d = d %>% mutate(param = colnames(m)) %>% select(param,post_mean:post_upper)
  
  d
}

strip_attrs = function(obj)
{
  attributes(obj) = NULL
  obj
}

strip_class = function(obj)
{
  attr(obj,"class") = NULL
  obj
}

nugget_cov   = function(d, sigma2=1) ifelse(d==0, sigma2, 0)
exp_cov      = function(d, sigma2=1, l=1) sigma2 * exp(-abs(d)*l)
sq_exp_cov   = function(d, sigma2=1, l=1) sigma2 * exp(-(abs(d)*l)^2)
pow_exp_cov  = function(d, sigma2=1, l=1, p=2) sigma2 * exp(-(abs(d)*l)^p)
rquad_cov    = function(d, sigma2=1, l=1, a=1) sigma2 * (1+d^2*l^2/a)^(-a)
periodic_cov = function(d, sigma2=1, l=1, p=1) sigma2 * exp(-2*l^2*sin(pi*d/p)^2)
matern_cov   = function(d, sigma2=1, l=1, nu=1/2) fields::Matern(d, alpha=l, nu=nu, phi=sigma2)
sphere_cov   = function(d, sigma2=1, l=1) ifelse(d > 1/l, 0, sigma2*(1 - 1.5*d*l + 0.5*(d*l)^3))

linear_cov   = function(x1, x2, sigma2_b=1, sigma2_v=1, c=0)
{
  if (!missing(x2))
  {
    sigma2_b + sigma2_v * (x1-c) * (x2-c)
  } else {
    expand.grid(t_i=x1, t_j=x1) %>%
    {linear_cov(.[[1]], .[[2]], sigma2_b=sigma2_b, sigma2_v=sigma2_v, c=c)} %>% 
      matrix(ncol=length(x1))
  }
}


  
# From https://github.com/hadley/dplyr/issues/271
bin = function(df, var, binwidth, origin = NULL) {
  n = nrow(df)

  var = as.character(substitute(var))
  x = df[[var]]

  if (is.null(origin)) {
    origin = min(x)
  }

  bin = (x - origin) %/% binwidth
  indices = unname(split(seq_len(n) - 1, bin))
  
  mid = origin + (0:(max(bin)+1)) * binwidth + binwidth/2
  
  df[["bin_mid"]] = mid[bin+1]
  
  attr(df, "indices") = indices
  attr(df, "drop") = FALSE
  attr(df, "group_sizes") = sapply(indices, length)
  attr(df, "biggest_group_size") = max(attr(df, "group_sizes"))
  attr(df, "labels") = data.frame(bin = seq_along(indices))
  attr(df, "vars") = list(quote(bin))
  class(df) = c("grouped_df", "tbl_df", "tbl", "data.frame")

  df
}



dist_long = function(d)
{
  d = as.matrix(d)
  d[upper.tri(d, diag = TRUE)] = NA
  
  data.frame(
    expand.grid(i=1:nrow(d), j=1:nrow(d)),
    c(d)
  ) %>%
    setNames(c("i","j","dist")) %>%
    filter(!is.na(dist))
}

emp_semivariogram = function(d, y, x, bin=FALSE, binwidth, range_max)
{
  y_col = as.character(substitute(y))
  x_col = as.character(substitute(x))
  
  d = d[[x_col]] %>%
    dist() %>% 
    dist_long() %>%
    mutate(y_i = d[[y_col]][i], y_j = d[[y_col]][j])
  
  
  if (bin)
  {
    d = d %>% bin(dist, binwidth = binwidth)
  } else {
    d = d %>% mutate(bin_mid = dist) %>% rowwise()
  }
  
  d = d %>%
    summarize(
      gamma = sum( (y_i - y_j)^2 / (2*n()) ),
      h = mean(bin_mid),
      n = n()
    )
  
  if (!missing(range_max))
    d = d %>% filter(h < range_max)
  
  d
}


rmvnorm = function(n, mu=rep(0, nrow(sigma)), sigma, diag_adjust = 1e-6)
{
  diag(sigma) = diag(sigma) + diag_adjust
  mu %*% matrix(1, ncol=n) + t(chol(sigma)) %*% matrix(rnorm(n*nrow(sigma)), ncol=n)
}
```


# Mauna Loa Exampel

## Atmospheric CO$_2$

```{r echo=FALSE, message=FALSE}
loa = data_frame(
  y = co2 %>% strip_attrs(),
  x = time(co2) %>% strip_attrs()
)

noaa = read_csv("noaa_mauna_loa.csv") %>%
  transmute(x = year+(month-1)/12, y=average) %>%
  filter(x > max(loa$x))

rbind(
  loa %>% mutate(Source="R"),
  noaa %>% mutate(Source="NOAA")
) %>%
ggplot(aes(x=x,y=y, color=Source)) + 
  geom_line() 
```

## ARIMA

```{r echo=FALSE}
tsdisplay(co2, points = FALSE)
```


## ARIMA(0,1,0)$\times$(0,0,0)

```{r echo=FALSE}
m1 = Arima(co2, order = c(0,1,0), seasonal = c(0,0,0))
tsdisplay(m1$residuals, points=FALSE)
m1$aicc
```

## ARIMA(0,1,0)$\times$(0,1,0)$_{12}$

```{r echo=FALSE}
m2 = Arima(co2, order = c(0,1,0), seasonal = c(0,1,0))
tsdisplay(m2$residuals, points=FALSE)
m2$aicc
```

## ARIMA(0,1,0)$\times$(0,1,1)$_{12}$

```{r echo=FALSE}
m3 = Arima(co2, order = c(0,1,0), seasonal = c(0,1,1))
tsdisplay(m3$residuals, points=FALSE)
m3$aicc
```

## ARIMA(0,1,1)$\times$(0,1,1)$_{12}$

```{r echo=FALSE}
m4 = Arima(co2, order = c(0,1,1), seasonal = c(0,1,1))
tsdisplay(m4$residuals, points=FALSE)
m4$aicc
```

## ARIMA(0,1,3)$\times$(0,1,1)$_{12}$

```{r echo=FALSE}
m5 = Arima(co2, order = c(0,1,3), seasonal = c(0,1,1))
tsdisplay(m5$residuals, points=FALSE)
m5$aicc
```

## auto.arima

```{r}
auto.arima(co2)
```


## Forecasting

```{r echo=FALSE}
m5_fore = forecast(m5,h = nrow(noaa))

plot(m5_fore )
lines(noaa, col='red', lwd=1.5)
```

## Forecasting (zoom)

```{r echo=FALSE}
plot(m5_fore, xlim=c(1998, 2017), ylim=c(360,410) )
lines(noaa, col='red', lwd=1.5)
```

## GP Model {.t}

Based on Rasmussen 5.4.3 (we are using slightly different data and parameterization)

$$ y \sim \mathcal{N}(\bm\mu,~ \bm\Sigma_1 + \bm\Sigma_2 + \bm\Sigma_3 + \bm\Sigma_4 + \sigma^2_5 \mathit{I}\,) $$

$$\{\bm\mu\}_i = \bar{y}$$


$$
\begin{aligned}
\{\bm\Sigma_1\}_{ij} &= \sigma^2_1 \exp\left(-(l_1 \cdot d_{ij})^2\right) \\
\{\bm\Sigma_2\}_{ij} &= \sigma^2_2 \exp\left(-(l_2 \cdot d_{ij})^2\right)\exp\left(-2 \, (l_3)^2  \sin^2(\pi \, d_{ij} / p)\right) \\
\{\bm\Sigma_3\}_{ij} &= \sigma^2_3 \left(1+\frac{(l_4 \cdot d_{ij})^2}{\alpha}\right)^{-\alpha} \\
\{\bm\Sigma_4\}_{ij} &= \sigma^2_4 \exp\left(-(l_5 \cdot d_{ij})^2\right)
\end{aligned}
$$

## JAGS Model

\scriptoutput
```{r echo=FALSE}
ml_model = "model{
  y ~ dmnorm(mu, inverse(Sigma))

  for (i in 1:(length(y)-1)) {
    for (j in (i+1):length(y)) {
      k1[i,j] <- sigma2[1] * exp(- pow(l[1] * d[i,j],2))
      k2[i,j] <- sigma2[2] * exp(- pow(l[2] * d[i,j],2) - 2 * pow(l[3] * sin(pi*d[i,j] / per), 2))
      k3[i,j] <- sigma2[3] * pow(1+pow(l[4] * d[i,j],2)/alpha, -alpha)
      k4[i,j] <- sigma2[4] * exp(- pow(l[5] * d[i,j],2))
      
      Sigma[i,j] <- k1[i,j] + k2[i,j] + k3[i,j] + k4[i,j]
      Sigma[j,i] <- Sigma[i,j]
    }
  }

  for (i in 1:length(y)) {
    Sigma[i,i] <- sigma2[1] + sigma2[2] + sigma2[3] + sigma2[4] + sigma2[5]
  }  

  for(i in 1:5){
    sigma2[i] ~ dt(0, 2.5, 1) T(0,)
    l[i] ~ dt(0, 2.5, 1) T(0,)
  }
  alpha ~ dt(0, 2.5, 1) T(0,)
}"
cat(ml_model,"\n")
```

```{r echo=FALSE}
if (!file.exists("init_model.Rdata"))
{
  m = jags.model(
    textConnection(gp_model), 
    data = list(
      mu = mean(loa$y) * rep(1, nrow(loa)),
      y = loa$y,
      #x = loa$x,
      d = dist(loa$x) %>% as.matrix(),
      per = 1,
      pi = pi
    ),
    inits = list(
      sigma2 = c(66, 2.4, 0.66, 0.18, 0.19)^2,
      l = 1/c(67, 90, 1.3, 1.2, 1.6/12),
      alpha = 0.78
    ),
    n.adapt=1000
  )
  save(m, file="init_model.Rdata")
} else {
  load("init_model.Rdata")
}

if (!file.exists("fit_model.Rdata"))
{
  gp_coda = coda.samples(
    m, variable.names=c("sigma2", "l", "alpha"),
    n.iter=1000
  )
  save(gp_coda, file="fit_model.Rdata")
} else {
  load("fit_model.Rdata")
}

sigma2 = get_coda_parameter(gp_coda, "sigma")
l      = get_coda_parameter(gp_coda, "^l")
alpha  = get_coda_parameter(gp_coda, "alpha")
```


<!--
## Model Fit

```{r echo=FALSE}
#y = loa$y
#x = loa$x
#
#n_post_samp = 1000
#
#calc_cov = function(d,i)
#{
#  S1 = sigma2[i,1] * exp(- (l[i,1] * d)^2)
#  S2 = sigma2[i,2] * exp(- (l[i,2] * d)^2 - 2 * (l[i,3] * sin(pi*d))^2)
#  S3 = sigma2[i,3] * (1+(l[i,4] * d)^2/alpha[i,1])^-alpha[i,1]
#  S4 = sigma2[i,4] * exp(- (l[i,5] * d)^2)
#  S5 = ifelse(abs(d)<1e-4, sigma2[i,5] + 1e-6, 0)
#  S = S1 + S2 + S3 + S4 + S5
#  
#  list(S1=S1, S2=S2, S3=S3, S4=S4, S5=S5, S=S)
#}
#
#x_pred = x-1/6
#y_pred = matrix(NA, nrow=n_post_samp, ncol=length(x_pred))
#colnames(y_pred) = paste0("Y_pred[", 1:length(x_pred), "]")
#
#mu = rep(mean(y), length(x))
#mu_pred = rep(mean(y), length(x_pred))
#
#dist_o = rdist(x)
#dist_p = rdist(x_pred)
#dist_op = rdist(x, x_pred)
#
#for(i in 1:nrow(sigma2))
#{
#  cov_o  = calc_cov(dist_o, i)$S
#  cov_p  = calc_cov(dist_p, i)$S
#  cov_op = calc_cov(dist_op, i)$S
# 
#  inv = solve(cov_o, cov_op)
#  cond_cov = cov_p - t(cov_op) %*% inv
#  diag(cond_cov) = diag(cond_cov) + 1e-5
#  cond_mu  = mu_pred + t(inv) %*% (y - mu)
#   
#  y_pred[i,] = cond_mu + t(chol(cond_cov)) %*% matrix(rnorm(length(x_pred)), ncol=1)
#}
#
#post_summary(y_pred) %>%
#  mutate(x = x_pred) %>%
#  ggplot(aes(x=x, y=post_mean)) +
#    geom_line(alpha=0.5, color='red') +
#    geom_ribbon(alpha=0.2, aes(ymin=post_lower, ymax=post_upper)) 
#    #geom_line(data=loa, aes(x=x,y=y), alpha=0.5)
```

-->


## Forecasting

```{r echo=FALSE}
y = loa$y
x = loa$x

n_post_samp = 1000

calc_cov = function(d,i)
{
  S1 = sigma2[i,1] * exp(- (l[i,1] * d)^2)
  S2 = sigma2[i,2] * exp(- (l[i,2] * d)^2 - 2 * (l[i,3] * sin(pi*d))^2)
  S3 = sigma2[i,3] * (1+(l[i,4] * d)^2/alpha[i,1])^-alpha[i,1]
  S4 = sigma2[i,4] * exp(- (l[i,5] * d)^2)
  S5 = ifelse(abs(d)<1e-4, sigma2[i,5] + 1e-6, 0)
  S = S1 + S2 + S3 + S4 + S5
  
  list(S1=S1, S2=S2, S3=S3, S4=S4, S5=S5, S=S)
}
```

```{r echo=FALSE}
if (!file.exists("ml_gp_fore.Rdata"))
{
  x_pred = noaa$x
  y_fore = matrix(NA, nrow=n_post_samp, ncol=length(x_pred))
  colnames(y_fore) = paste0("Y_pred[", 1:length(x_pred), "]")
  
  mu = rep(mean(y), length(x))
  mu_pred = rep(mean(y), length(x_pred))
  
  dist_o = rdist(x)
  dist_p = rdist(x_pred)
  dist_op = rdist(x, x_pred)
  
  for(i in 1:nrow(sigma2))
  {
    cov_o  = calc_cov(dist_o, i)$S
    cov_p  = calc_cov(dist_p, i)$S
    cov_op = calc_cov(dist_op, i)$S
    
    inv = solve(cov_o, cov_op)
    cond_cov = cov_p - t(cov_op) %*% inv
    diag(cond_cov) = diag(cond_cov) + 1e-5
    cond_mu  = mu_pred + t(inv) %*% (y - mu)
    
    y_fore[i,] = cond_mu + t(chol(cond_cov)) %*% matrix(rnorm(length(x_pred)), ncol=1)
  }
  
  save(y_fore, file="ml_gp_fore.Rdata")
} else {
  load(file="ml_gp_fore.Rdata")
}
```

```{r echo=FALSE}
summ_80 = post_summary(y_fore, ci_width = 0.8) %>% mutate(x = noaa[["x"]])
summ_95 = post_summary(y_fore, ci_width = 0.95) %>% mutate(x = noaa[["x"]])

ggplot(summ_95, aes(x=x, y=post_mean)) +
  geom_ribbon(data=summ_95, aes(ymin=post_lower, ymax=post_upper), fill="#D3D2D6") +
  geom_ribbon(data=summ_80, aes(ymin=post_lower, ymax=post_upper), fill="#A3A5C3") +
  geom_line(color="#1007FC") +
  geom_line(data=loa, aes(x=x,y=y), alpha=0.5) +
  geom_line(data=noaa, aes(x=x,y=y), alpha=0.5, col='red')
```

## Forecasting (zoom)

```{r echo=FALSE}
ggplot(summ_95, aes(x=x, y=post_mean)) +
  geom_ribbon(data=summ_95, aes(ymin=post_lower, ymax=post_upper), fill="#D3D2D6") +
  geom_ribbon(data=summ_80, aes(ymin=post_lower, ymax=post_upper), fill="#A3A5C3") +
  geom_line(color="#1007FC", size=0.8) +
  geom_line(data=loa, aes(x=x,y=y), alpha=0.5, size=0.8) +
  geom_line(data=noaa, aes(x=x,y=y), alpha=0.5, col='red', size=0.8) +
  xlim(1998, 2017) + ylim(360,410)
```


## Forecasting RMSE

```{r echo=FALSE}
noaa_pred_arima = data_frame(
  y_hat = m5_fore$mean %>% strip_attrs(),
  x = time(m5_fore$mean) %>% strip_attrs()
) %>% 
  mutate(y = noaa[["y"]])

noaa_pred_gp = summ_95 %>%
  select(y_hat = post_mean, x) %>%
  mutate(y = noaa[["y"]])

filters = c(2003, 2008, 2013, 2018)

noaa_rmse = function(year, d) 
{
  d %>% 
    filter(x < year) %>% 
    {(.$y - .$y_hat)^2} %>% 
    mean() %>% 
    sqrt() %>% 
    round(3)
}

data_frame(
  "dates"        = c("Jan 1998 - Jan 2003", "Jan 1998 - Jan 2008", "Jan 1998 - Jan 2013", "Jan 1998 - Mar 2017"),
  "RMSE (arima)" = map_dbl(filters, noaa_rmse, d=noaa_pred_arima),
  "RMSE (gp)"    = map_dbl(filters, noaa_rmse, d=noaa_pred_gp)
) %>%
  knitr::kable()
```

 
## Rewriting the GP likelihood {.t}

From last time, remember that we can view our GP in the following ways,

$$ y \sim \mathcal{N}(\bm\mu,~ \bm\Sigma_1 + \bm\Sigma_2 + \bm\Sigma_3 + \bm\Sigma_4 + \sigma^2_5 \mathit{I}\,) $$

but we can also think of $y$ as being the deterministic sum of 5 independent GPs

$$ y = \mu + w_1(\bm{x}) + w_2(\bm{x}) + w_3(\bm{x}) + w_4(\bm{x}) + w_5(\bm{x}) $$
where 
$$
\begin{aligned}
w_1(\bm{x}) &\sim \mathcal{N}(0, \bm\Sigma_1) \\
w_2(\bm{x}) &\sim \mathcal{N}(0, \bm\Sigma_2) \\
w_3(\bm{x}) &\sim \mathcal{N}(0, \bm\Sigma_3) \\
w_4(\bm{x}) &\sim \mathcal{N}(0, \bm\Sigma_4) \\
w_5(\bm{x}) &\sim \mathcal{N}(0, \sigma^2_5 \mathit{I}\,)
\end{aligned}
$$

## Decomposition of Covariance Components {.t}

\footnotesize
$$ \begin{bmatrix} 
w_1(\bm{x}) \\
w_1(\bm{x}^\star) \\
w_2(\bm{x}) \\
\bm{y}
\end{bmatrix} 
\sim \mathcal{N} \left(
\begin{bmatrix}
0 \\ 0\\ 0\\ \bm{\mu}
\end{bmatrix},~
\begin{bmatrix}
\Sigma_1 & \Sigma_1^\star & 0 & \Sigma_1 \\
{\Sigma_1^\star}^t & \Sigma_1^{\star\star} & 0 & \Sigma_1^\star \\
0 & 0 & \Sigma_2 & \Sigma_2 \\
\Sigma_1 & \Sigma_1^\star & \Sigma_2 & \sum_{i=1}^5 \Sigma_i
\end{bmatrix}
\right)
$$

\normalsize


therefore

$$ w_1(\bm{x}^\star) ~|~ \bm{y},\bm\mu,\bm\theta \sim \mathcal{N}(\mu_{cond},~ \Sigma_{cond}) $$

$$ \mu_{cond} = 0 + \Sigma_1^\star (\Sigma_1 + \Sigma_2 + \Sigma_3 + \Sigma_4 + \Sigma_5)^{-1}(\bm{y}-\bm\mu) $$
$$ \Sigma_{cond} = \Sigma_1^{\star\star} - \Sigma_1^\star (\Sigma_1 + \Sigma_2 + \Sigma_3 + \Sigma_4 + \Sigma_5)^{-1} {\Sigma_1^\star}^t $$



## Forecasting Components

```{r echo=FALSE}
if (!file.exists("ml_gp_fore_comp.Rdata"))
{
  x_pred = noaa$x
  y_comp = lapply(1:4, function(i) matrix(NA, nrow=n_post_samp, ncol=length(x_pred)))
    
  mu = rep(mean(y), length(x))
  mu_pred = rep(0, length(x_pred))
  
  dist_o = rdist(x)
  dist_p = rdist(x_pred)
  dist_op = rdist(x, x_pred)
  
  p = progress_estimated(n_post_samp)
  for(i in 1:nrow(sigma2))
  {
    cov_o  = calc_cov(dist_o, i)$S
    inv = solve(cov_o)
    
    for(j in 1:4)
    {
      cov_name = paste0("S",j)
      cov_p  = calc_cov(dist_p, i)[[cov_name]]
      cov_op = calc_cov(dist_op, i)[[cov_name]]
      
      cond_cov = cov_p - t(cov_op) %*% inv %*% cov_op
      diag(cond_cov) = diag(cond_cov) + 1e-4
      cond_mu  = mu_pred + t(cov_op) %*% inv %*% (y - mu)
    
      y_comp[[j]][i,] = cond_mu + t(chol(cond_cov)) %*% matrix(rnorm(length(x_pred)), ncol=1)
    }
    p$tick()$print()
  }
  
  save(y_comp, file="ml_gp_fore_comp.Rdata")
} else {
  load(file="ml_gp_fore_comp.Rdata")
}
```

```{r echo=FALSE}
lapply(
  1:4, 
  function(i) 
    post_summary(y_comp[[i]]) %>% 
      mutate(cov=paste0("Sigma_",i), x=noaa[["x"]])
) %>% 
  bind_rows() %>%
  ggplot(aes(x=x, y=post_mean, col=cov, fill=cov)) +
    geom_ribbon(aes(ymin=post_lower, ymax=post_upper, color=NULL), alpha=0.2) +
    geom_line() +
    facet_wrap(~cov, scale="free_y")
```

## Fit Components

```{r echo=FALSE}
if (!file.exists("ml_gp_fit_comp.Rdata"))
{
  x_pred = loa$x
  y_fit_comp = lapply(1:4, function(i) matrix(NA, nrow=n_post_samp, ncol=length(x_pred)))
    
  mu = rep(mean(y), length(x))
  mu_pred = rep(0, length(x_pred))
  
  dist_o = rdist(x)
  dist_p = rdist(x_pred)
  dist_op = rdist(x, x_pred)
  
  p = progress_estimated(n_post_samp)
  for(i in 1:nrow(sigma2))
  {
    cov_o  = calc_cov(dist_o, i)$S
    inv = solve(cov_o)
    
    for(j in 1:4)
    {
      cov_name = paste0("S",j)
      cov_p  = calc_cov(dist_p, i)[[cov_name]]
      cov_op = calc_cov(dist_op, i)[[cov_name]]
      
      cond_cov = cov_p - t(cov_op) %*% inv %*% cov_op
      diag(cond_cov) = diag(cond_cov) + 1e-4
      cond_mu  = mu_pred + t(cov_op) %*% inv %*% (y - mu)
    
      y_fit_comp[[j]][i,] = cond_mu + t(chol(cond_cov)) %*% matrix(rnorm(length(x_pred)), ncol=1)
    }
    p$tick()$print()
  }
  
  save(y_fit_comp, file="ml_gp_fit_comp.Rdata")
} else {
  load(file="ml_gp_fit_comp.Rdata")
}
```

```{r echo=FALSE}
lapply(
  1:4, 
  function(i) 
    post_summary(y_fit_comp[[i]]) %>% 
      mutate(cov=paste0("Sigma_",i), x=loa[["x"]])
) %>% 
  bind_rows() %>%
  ggplot(aes(x=x, y=post_mean, col=cov, fill=cov)) +
    geom_ribbon(aes(ymin=post_lower, ymax=post_upper, color=NULL), alpha=0.2) +
    geom_line() +
    facet_wrap(~cov, scale="free_y")
```


# GPs and Logistic Regression

## Logistic Regression {.t}
 
A typical logistic regression problem uses the following model,

$$y_i \sim \text{Bern}(p_i)$$
$$\begin{aligned}
\text{logit}(p_i) 
  &= \bm{X}\,\bm{\beta} \\
  &= \beta_0 + \beta_1 \, x_{i1} + \cdots + \beta_k \, x_{ik}
\end{aligned}$$

. . .

there is no reason that the linear equation above can't contain thing like random effects or GPs

$$\begin{aligned}
y_i &\sim \text{Bern}(p_i) \\
\text{logit}(p_i) 
  &= \bm{X}\,\bm{\beta} + w(\bm{x}) \\
\end{aligned}$$
where
$$ w(\bm{x}) \sim \mathcal{N}(0,\Sigma)$$

## A toy example

```{r echo=FALSE}
inv_logit = function(x) exp(x) / (1+exp(x))

x = c( abs(rnorm(30, 0, 0.1)), rnorm(30, 0.5, 0.05), 1-abs(rnorm(30, 0, 0.1)))
eta = 2*sin(2.1*pi*(x-0.25)) + rnorm(90, sd=0.1)
y = rbinom(length(eta), size=1, prob = inv_logit(eta))

d = data_frame(
  x=x, y=y
)

ggplot(d, aes(x=x, y=y)) +
  geom_jitter(height = 0.05)
```


## Jags Model

\footnoteoutput
```{r echo=FALSE}
logistic_model = "model{
  for(i in 1:N) {
    y[i] ~ dbern(p[i])
    logit(p[i]) <- eta[i]
  }
  eta ~ dmnorm(rep(0,N), inverse(Sigma))

  for (i in 1:(length(y)-1)) {
    for (j in (i+1):length(y)) {
      Sigma[i,j] <- sigma2 * exp(- pow(l * d[i,j],2))
      Sigma[j,i] <- Sigma[i,j]
    }
  }

  for (i in 1:length(y)) {
    Sigma[i,i] <- sigma2 + 1e-06
  }  

  sigma2 ~ dt(0, 2.5, 1) T(0,)
  l ~ dunif(sqrt(3),100)
}"
cat(logistic_model,"\n")
```

## Model Results

```{r echo=FALSE}
if (!file.exists("logistic_model.Rdata"))
{
  m_log = jags.model(
    textConnection(logistic_model), 
    data = list(
      d = dist(d$x) %>% as.matrix(),
      y = d$y,
      x = d$x,
      N = nrow(d)
    ),
    n.adapt=5000
  )

  update(m_log, n.iter=10000)#, progress.bar="none")
  
  log_coda = coda.samples(
    m_log, variable.names=c("sigma2", "l", "eta"),
    n.iter=10000, thin=10
  )
  save(m_log, logistic_model, log_coda, file="logistic_model.Rdata")
} else {
  load("logistic_model.Rdata")
}

cov_params = get_coda_parameter(log_coda, "sigma|l")
eta = get_coda_parameter(log_coda, "eta")
```

```{r echo=FALSE}
eta_post = eta %>%
    post_summary() %>%
    mutate(x = d[["x"]], param="eta")

p_post = eta %>%
    inv_logit() %>%
    post_summary() %>%
    mutate(x = d[["x"]], param="p")

rbind(eta_post, p_post) %>%
  ggplot(aes(x=x, y=post_mean, color=param, fill=param)) +
    geom_ribbon(aes(ymin=post_lower, ymax=post_upper), alpha=0.2, color=NA) +
    geom_line() +
    facet_wrap(~param, scale="free_y")
```