
---
title: "Lecture 11" 
subtitle: "Seasonal ARIMA"
author: "Colin Rundel"
date: "02/22/2017"
fontsize: 11pt
output: 
  beamer_presentation:
    theme: metropolis
    highlight: pygments
    fig_width: 8
    fig_height: 6
    fig_caption: false
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: ../settings.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning=FALSE, message=FALSE)
options(width=70)

set.seed(20170222)

library(magrittr)
library(dplyr)
library(modelr)
library(ggplot2)
library(tidyr)
library(rjags)
library(stringr)
library(gridExtra)
library(readr)
library(purrr)
library(forcats)
library(forecast)
library(astsa)

theme_set(
  theme_bw()  
)

get_coda_parameter = function(coda, pattern)
{
  w = coda[[1]] %>% colnames() %>% str_detect(pattern)
  coda[[1]][,w,drop=FALSE]
}

post_summary = function(m, ci_width=0.95)
{
  d = data_frame(
    post_mean  = apply(m, 2, mean),
    post_lower = apply(m, 2, quantile, probs=(1-ci_width)/2),
    post_upper = apply(m, 2, quantile, probs=1 - (1-ci_width)/2)
  )
  rownames(d) = colnames(m)
  d
}

strip_attrs = function(obj)
{
  attributes(obj) = NULL
  obj
}

strip_class = function(obj)
{
  attr(obj,"class") = NULL
  obj
}
```

# Seasonal Models

## Australian Wine Sales Example (Lecture 6)

Australian total wine sales by wine makers in bottles <= 1 litre. Jan 1980 – Aug 1994.

```{r echo=FALSE}
tsdisplay(wineind, points = FALSE)
```

## Differencing

```{r echo=FALSE}
tsdisplay(diff(wineind), points = FALSE)
```

## Seasonal ARIMA {.t}

We can extend the existing ARIMA model to handle these higher order lags (without having to include all of the intervening lags).

\vspace{3mm}

Seasonal $\text{ARIMA}\,(p,d,q) \times (P,D,Q)_s$:
$$ \Phi_P(L^s) \, \phi_p(L) \, \Delta_s^D \, \Delta^d \, y_t = \delta + \Theta_Q(L^s) \, \theta_q(L) \, w_t$$
\pause where
$$
\begin{aligned}
\phi_p(L) &= 1-\phi_1 L - \phi_2 L^2 - \ldots - \phi_p L^p\\
\theta_q(L) &= 1+\theta_1 L + \theta_2 L^2 + \ldots + \theta_p L^q \\
\Delta^d &= (1-L)^d\\
\\
\Phi_P(L^s) &= 1-\Phi_1 L^s - \Phi_2 L^{2s} - \ldots - \Phi_P L^{Ps} \\
\Theta_Q(L^s) &= 1+\Theta_1 L + \Theta_2 L^{2s} + \ldots + \theta_p L^{Qs} \\
\Delta_s^D &= (1-L^s)^D\\
\end{aligned}
$$

## Seasonal ARIMA for \texttt{wineind}

Lets consider an $\text{ARIMA}(0,0,0) \times (1,0,0)_{12}$:
$$
\begin{aligned}
(1-\Phi_1 L^{12}) \, y_t = \delta + w_t \\
y_t = \Phi_1 y_{t-12} + \delta + w_t
\end{aligned}
$$
\vspace{2mm}

```{r}
(m1 = Arima(wineind, seasonal=list(order=c(1,0,0), period=12)))
```

## Fitted model

```{r echo=FALSE, fig.height=4}
cbPalette = c("#000000", "#ff3333", "#92c5de")

model = "Model 1 - ARIMA (0,0,0) x (0,1,0)[12]"
rmse = (wineind-m1$fitted)^2 %>% mean() %>% sqrt() %>% round(2) %>% paste0("      [RMSE: ", . ,"]")

m1 %>%
  {data_frame(
    wineind = wineind %>% strip_attrs(),
    model   = .$fitted %>% strip_attrs(),
    time    = time(.$fitted) %>% strip_attrs()
  )} %>% 
  gather(var, sales, -time) %>%
  mutate(var = as_factor(var)) %>%
  ggplot(aes(x=time, y=sales, color=var)) + 
    geom_line(alpha=0.75, size=0.8) +
    scale_colour_manual(values=cbPalette, name="") +
    labs(title=paste(model, rmse))
```

## Residuals

```{r echo=FALSE}
tsdisplay(m1$residuals)
```


## Model 2

$\text{ARIMA}(0,0,0) \times (0,1,1)_{12}$:
$$
\begin{aligned}
(1-L^{12}) y_t = \delta + (1+\Theta_1 L^{12})  w_t \\
y_t-y_{t-12} = \delta + w_t + \Theta_1 w_{t-12} \\
y_t = \delta + y_{t-12} + w_t + \Theta_1 w_{t-12}
\end{aligned}
$$


```{r}
(m2 = Arima(wineind, order=c(0,0,0), seasonal=list(order=c(0,1,1), period=12)))
```

## Fitted model

```{r echo=FALSE, fig.height=4}
model = "Model 2 - ARIMA (0,0,0) x (0,1,1)[12]"
rmse = (wineind-m2$fitted)^2 %>% mean() %>% sqrt() %>% round(2) %>% paste0("      [RMSE: ", . ,"]")

m2 %>%
  {data_frame(
    wineind = wineind %>% strip_attrs(),
    model   = .$fitted %>% strip_attrs(),
    time    = time(.$fitted) %>% strip_attrs()
  )} %>% 
  gather(var, sales, -time) %>%
  mutate(var = as_factor(var)) %>%
  ggplot(aes(x=time, y=sales, color=var)) + 
    geom_line(alpha=0.75, size=0.8) +
    scale_colour_manual(values=cbPalette, name="") +
    labs(title=paste(model, rmse))
```

## Residuals

```{r echo=FALSE}
tsdisplay(m2$residuals)
```


## Model 3

$\text{ARIMA}(3,0,0) \times (0,1,1)_{12}$
$$
\begin{aligned}
(1-\phi_1 L - \phi_2 L^2 - \phi_3 L^3) \, (1-L^{12}) y_t = \delta + (1 + \Theta_1 L)w_t \\
(1-\phi_1 L - \phi_2 L^2 - \phi_3 L^3) \, (y_t-y_{t-12}) = \delta + w_t + w_{t-12} \\
y_t = \delta + \sum_{i=1}^3 \phi_i y_{t-1}  + y_{t-12}  - \sum_{i=1}^3 \phi_i y_{t-12-i} + w_t + w_{t-12}
\end{aligned}
$$


```{r}
(m3 = Arima(wineind, order=c(3,0,0), 
            seasonal=list(order=c(0,1,1), period=12)))
```

## Fitted model

```{r echo=FALSE, fig.height=4}
model = "Model 3 - ARIMA (3,0,0) x (0,1,1)[12]"
rmse = (wineind-m3$fitted)^2 %>% mean() %>% sqrt() %>% round(2) %>% paste0("      [RMSE: ", . ,"]")

m3 %>%
  {data_frame(
    wineind = wineind %>% strip_attrs(),
    model   = .$fitted %>% strip_attrs(),
    time    = time(.$fitted) %>% strip_attrs()
  )} %>% 
  gather(var, sales, -time) %>%
  mutate(var = as_factor(var)) %>%
  ggplot(aes(x=time, y=sales, color=var)) + 
    geom_line(alpha=0.75, size=0.8) +
    scale_colour_manual(values=cbPalette, name="") +
    labs(title=paste(model, rmse))
```

## Model - Residuals

```{r echo=FALSE}
tsdisplay(m3$residuals)
```

# Federal Reserve Board Production Index

## \texttt{prodn} from the \texttt{astsa} package

Monthly Federal Reserve Board Production Index (1948-1978)

```{r echo=FALSE}
tsdisplay(prodn, points = FALSE)
```

## Differencing

Based on the ACF it seems like standard differencing may be required

```{r echo=FALSE}
tsdisplay(diff(prodn))
```

## Differencing + Seasonal Differencing

Additional seasonal differencing also seems warranted

```{r}
(fr_m1 = Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,0,0), period=12)))
```

```{r}
(fr_m2 = Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,1,0), period=12)))
```


## Residuals

```{r echo=FALSE}
tsdisplay(fr_m2$residuals, points=FALSE, lag.max=36)
```

## Adding Seasonal MA

\scriptoutput

```{r}
(fr_m3.1 = Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,1,1), period=12)))
```

```{r}
(fr_m3.2 = Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,1,2), period=12)))
```

## Adding Seasonal MA (cont.) {.t}

\scriptoutput 

```{r}
(fr_m3.3 = Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,1,3), period=12)))
```


## Residuals - Model 3.3

```{r echo=FALSE}
tsdisplay(fr_m3.3$residuals, points=FALSE, lag.max = 36)
```


## Adding AR

\scriptoutput

```{r}
(fr_m4.1 = Arima(prodn, order = c(1,1,0), 
            seasonal = list(order=c(0,1,3), period=12)))
```

```{r}
(fr_m4.2 = Arima(prodn, order = c(2,1,0), 
            seasonal = list(order=c(0,1,3), period=12)))
```

## Residuals - Model 4.1

```{r echo=FALSE}
tsdisplay(fr_m4.1$residuals, points=FALSE, lag.max = 36)
```

## Residuals - Model 4.2

```{r echo=FALSE}
tsdisplay(fr_m4.2$residuals, points=FALSE, lag.max = 36)
```

## Model Fit

```{r echo=FALSE, fig.height=4}
model = "Model 4.1 - ARIMA (1,1,0) x (0,1,3)[12]"
rmse = (prodn-fr_m4.1$fitted)^2 %>% mean() %>% sqrt() %>% round(3) %>% paste0("      [RMSE: ", . ,"]")

fr_m4.1 %>%
  {data_frame(
    prodn = prodn %>% strip_attrs(),
    model = .$fitted %>% strip_attrs(),
    time  = time(.$fitted) %>% strip_attrs()
  )} %>% 
  gather(var, sales, -time) %>%
  mutate(var = as_factor(var)) %>%
  ggplot(aes(x=time, y=sales, color=var)) + 
    geom_line(alpha=0.75, size=0.8) +
    scale_colour_manual(values=cbPalette, name="") +
    labs(title=paste(model, rmse))
```

## Model Forecast

```{r echo=FALSE}
forecast(fr_m4.1) %>% plot()
```

## Model Forecast (cont.)

```{r echo=FALSE}
forecast(fr_m4.1) %>% plot(xlim=c(1975,1982))
```

## Model Forecast (cont.)

```{r echo=FALSE}
forecast(fr_m4.1, 120) %>% plot()
```

## Model Forecast (cont.)

```{r echo=FALSE}
forecast(fr_m4.1, 120) %>% plot(xlim=c(1975,1990))
```


## Exercise - Cortecosteroid Drug Sales

Monthly cortecosteroid drug sales in Australia from 1992 to 2008.

```{r}
library(fpp)
tsdisplay(h02,points=FALSE)
```

## Hint

```{r}
ts.intersect(h02, log(h02)) %>% plot()
```
